Retinotopic coding organizes the interaction between internally and externally oriented brain networks

Version 1

Abstract

The human brain seamlessly integrates internally generated thoughts with incoming sensory information, yet the large-scale networks that support these functions -- the (Default Network, DN) and external (Dorsal Attention Network, dATN) -- are traditionally viewed as functionally antagonistic. This raises a crucial question: how does the brain integrate information across these seemingly opposed systems? Here, using densely sampled 7T fMRI, individualized resting-state parcellations, and voxel-wise population-receptive-field mapping, we show that these internal/external networks are more interlocked than previously thought. Although spontaneous DN and dATN activity during rest is uncorrelated at the network level, functional coupling across networks is shaped by the latent visual field preferences of individual voxels in each network, as measured during independent retinotopic mapping. Voxels that share visual field preferences exhibit stronger spontaneous coupling than those with divergent preferences. These retinotopically-specific interactions are bivalent: DN voxels with negative (suppressive) visual response amplitudes are anticorrelated with matched (positive) dATN voxels, while those with positive response amplitudes are positively correlated. Thus, distinct subpopulations of visually-tuned DN voxels participate in spatially-specific interactions with the dATN. Further, retinotopic coding is intrinsic to the DN, persisting even during periods of elevated top-down drive from the DN to the dATN. These findings challenge the prevailing view of global DN-dATN antagonism, revealing a latent, voxel-level architecture of retinotopically-grounded interactions. Taken together, our results suggest that retinotopic coding underpins the dynamic coordination of perception and thought in the human brain.

Main Text

A fundamental goal of neuroscience is to understand how the coordinated activity of distributed brain networks gives rise to cognition (1–9). Central to this aim is understanding principles that govern interactions among networks with distinct computational roles. In particular, it is unclear how networks involved in externally-oriented attention (e.g., processing sensory input) interact with those that support internally-oriented attention (e.g., introspection and memory) (10–14). Understanding this interplay is crucial, because the dynamic exchange between internal and external representations underlies how we perceive, remember, and navigate our world -- yet the ‘common language’ that enables communication between these systems remains unclear.

Historically, the Default Network (DN) and Dorsal Attention Network (dATN)—prototypical networks for internal and external attention, respectively—have been viewed as functionally competitive (10–12, 14, 15). Classic neuroimaging studies show that externally oriented tasks activate the dATN and deactivate the DN (10, 13, 14, 16–18), while internally oriented tasks produce the opposite pattern (6, 19–23). This dissociation is also observed at rest, where DN and dATN activity is typically independent or anti-correlated (11, 12, 24, 25), depending on the application of global signal regression (11, 26). Such findings have led to the prevailing view that DN/dATN antagonism supports selective attention to either internal or external information, minimizing interference between perception and memory (10, 12, 27, 28). Yet, this framework leaves open a fundamental question: if these networks are functionally dissociated, how does the brain integrate perceptual and mnemonic information when both are required – for example during anticipatory saccades or memory-guided attention?

Recent studies of the voxel-level coding properties within these large-scale networks offer insight into this question. Although the DN is traditionally associated with abstract or semantic coding (2, 28–30), emerging evidence suggests that it also exhibits a neural code typically associated with externally-oriented visual processing: coding of retinotopic stimulus position (31–34). Notably, retinotopic coding in the DN differs from that observed in traditional visually responsive areas, including the dATN. While visual stimulation elicits position-dependent increases in neural activity in traditional visual areas, many DN voxels often show position-dependent decreases in activity(31, 33, 34). This voxel-level pattern suggests that the DN’s global deactivation during externally-oriented tasks may not reflect a simple disengagement from external input, but rather a structured, spatially-tuned response to visual stimulation.

What purpose might the DN’s voxel-level retinotopic code serve? We and others have proposed that the visual code in the DN might play a functional role in structuring mnemonic-perceptual interactions (31, 32, 34, 35). Recent work provides critical support for this theory, revealing that scene-selective memory regions near the DN (36) exhibit voxel-level retinotopic opponency with scene-selective visual areas (34). Specifically, the inverted retinotopic code in mnemonic regions is functionally tethered to the positive code in perceptual regions: during perception, visual stimuli at specific field locations activate perceptual voxels while suppressing memory voxels tuned to the same location, a pattern that reverses during memory retrieval (34). These results indicate that visual field location can coordinate dynamic interactions between specialized perceptual and mnemonic areas at the cortical apex, raising the intriguing hypothesis that retinotopic grounding might support perceptual-mnemonic interactions more broadly across the brain. Yet, whether retinotopic coding serves as a more general organizing principle for interactions across large-scale networks remains unknown.

Here, we tested the hypothesis that global interactions across large-scale internally- and externally-oriented brain networks are structured at the voxel level by a shared retinotopic code35. This framework challenges the traditional view of internally- and externally-oriented brain networks as globally antagonistic, proposing instead that voxel-level retinotopic tuning enables functional coupling and shared information processing between them. To evaluate this hypothesis, we leveraged a high-resolution 7T fMRI dataset(37), individualized network identification(3, 4, 8), and voxel-wise modeling(38) to test three core predictions. First, voxel-wise retinotopic coding should be robustly present within individuals’ DN and dATN networks, but with divergent tuning polarities: positive retinotopic responses should dominate the dATN, while DN should contains a significant proportion of retinotopic voxels with negative response amplitudes to visual stimulation. Second, we predicted that spontaneous DN–dATN interactions during resting-state fMRI would be structured by the retinotopic properties of individual voxels. Specifically, functional coupling should be stronger between voxels that share visual field preferences across networks, revealing that retinotopic position-specific responses form a latent architecture to scaffold cross-network interactions even in the absence of overt task demands. Third, we predicted that these retinotopically grounded interactions would support both top-down and bottom-up coupling between the DN and dATN, suggesting that shared spatial tuning enables dynamic, bidirectional interactions. Together, these findings would position retinotopy as a unifying framework for brain-wide information processing and for the coordination of internal and external attentional dynamics.

Retinotopic coding in internally and externally oriented networks

To investigate the role of retinotopic coding in structuring interactions between internally- and externally-oriented brain networks, we analyzed 7T fMRI data from the Natural Scenes Dataset(37), acquired from the same densely-sampled participants during both resting-state and retinotopic mapping sessions (Fig. 1A-B).

Inversion of retinotopic coding between externally- and internally-oriented brain networks.

A. Population receptive field (pRF) modeling with fMRI. We established visual field preferences for each voxel using pRF modelling(38). Voxels with positive BOLD responses to the visual stimulus are classified as positive pRFs (+pRFs), and those with negative BOLD responses are classified as negative pRFs (-pRFs). B. Individualized resting-state network parcellation. Resting-state fMRI was collected in all participants (N=7; 34-102 minutes per participant) and used to derive individualized cortical network parcellations. Parcellations were generated using the multi-session hierarchical Bayesian modelling approach with the Yeo 15 HCP atlas as a prior(8, 42). C. PRF model variance explained is greater when fit to the pRF mapping task versus resting-state fMRI data for negative and positive pRFs in the DN and dATN. Histograms show voxel-wise variance explained for all pRFs fit to rest (dark) and the pRF modeling task fMRI data (light). All distributions were significantly shifted rightward for data fit to the pRF modeling task (dATN pRFs: K-S test: D-=0.655, p<0.001; Negative DN pRFs: D-=0.232, p<0.001; Positive DN pRFs: D-=0.124, p<0.001). Dashed lines indicate the 95th percentile of R2 within the null distribution used; we chose the threshold for visually-defined voxels for subsequent analyses (R2 > 0.14) to exceed these values. D. DN and dATN pRF amplitude and center location estimates are reliable. To assess reliability of pRF amplitude estimates, we compared voxel-wise signed amplitude by computing the dice coefficients of binarized pRF amplitude maps (i.e., positive or negative pRFs) between pairs of pRF modelling runs. To estimate center position, we calculated the distance between pairs of pRF centers iteratively between pRF modeling runs. For both metrics, chance was estimated by performing the same computation on a random sampling of voxels drawn from both pRFs and non-pRFs (5000 iterations for each metric). In all subjects, for dATN positive and DN positive and negative pRFs, pRF amplitude and center position exceeded what would be expected by chance (z-scored Dice/distance values > 2.9). Although only dATN positive pRFs were included in subsequent analyses, the dATN pRF amplitude was not reliable above chance in two participants; this suggests that negative amplitude pRFs are not stable in all cortical networks. E. The internally-oriented DN contains a larger proportion of negative pRFs than the externally-oriented dATN. Proportion of negative pRFs (of total pRFs that survive thresholding in that network) in all cortical networks are shown. Inset. DN-A and DN-B contain more negative pRFs than the dATN (t(6)=7.94, p<0.001), and there is no difference in the proportion of negative pRFs among sub-networks. F. DN and dATN pRFs networks show contralateral bias. Coverage plots across participants for positive dATN pRFs and positive and negative DN pRFs. Bar plots show the difference in coverage between contra-versus ipsi-lateral visual fields for each pRF type. Bar plots show mean across participants, connected points represent individual participant data.

We first identified individualized large-scale networks for each participant using their resting-state fMRI data (8, 39) (Fig. 1B, S1). Resting-state data were preprocessed using independent components analysis (ICA) with manual noise component selection(40, 41), and no global signal regression was performed(26). Networks were identified using a multi-session hierarchical Bayesian modeling approach(8, 39, 42), which leverages repeated data runs to estimate within participant variation and derive stable parcellations for 15 cortical networks, including the Default Networks A-B and the Dorsal Attention Networks A-B (Fig. 1B). Unless otherwise noted, we combined the Default Networks A-B and the Dorsal Attention Networks A-B to constitute our internal and externally oriented networks, respectively (hereafter, DN and dATN).

Next, we determined the visual field preferences of all voxels within these individualized networks using a voxel-wise population receptive field (pRF) modeling approach(38, 43) to fit each participant’s BOLD activity during retinotopic mapping(38) (Fig 1A, S2). We considered any voxel with >14% variance explained by our pRF model to be exhibiting a retinotopic code, and hereafter we refer to these voxels as “pRFs”. This threshold was determined by defining the noise floor of our retinotopic model in the DN and dATN by comparing task-derived model fits to a control null distribution derived from resting-state fMRI data (Fig. 1C; distribution of voxel-wise variance explained by task versus null, K-S tests: D-s>0.124, ps<0.001). Two pRF features were of primary interest: 1) center position, which estimates the preferred location in visual space that drives a given voxel’s response (x,y), and 2) amplitude of the voxel’s response (positive or negative) (Fig. 1A). Importantly, both features were highly reproducible across runs of pRF data in the dATN and DN in all subjects (Fig. 1D).

Next, we compared the retinotopic properties of voxels within the DN and dATN(11, 12, 18). Across participants, we observed a larger proportion of retinotopic voxels in the dATN as compared to the DN (43.36% of dATN voxels, 9.42% of DN voxels) consistent with their role as externally- and internally-oriented networks, respectively. Further, the amplitude of retinotopic responses (i.e. whether a stimulus in their preferred visual field location evoked a positive or a negative BOLD response) differed significantly between these two networks (Fig 1E). Consistent with prior reports, negative voxel-wise response amplitudes were clustered within the DN and in similar locations across participants (31, 34), evidenced by a consensus map of pRF sign (Fig. S3). On average, more than half (58.77%) of all pRFs in the DNs were inverted (i.e., had a negative BOLD response to visual stimulation in their population receptive field, -pRFs). In contrast, just 7.57% of the pRFs in the dATN were -pRFs (i.e., the vast majority of the voxels had positive BOLD responses to visual stimulation in their receptive field, +pRFs). The difference in negative pRF proportion between the dATNs and DNs was significant (t(6)=7.94, p<0.001; Fig. 1E, inset), and there was no difference among the sub-networks of each system (dATNs: t(6)=0.24, p=0.82; DNs: t(6)=0.89, p=0.41). This distinction is particularly remarkable given the proximity of dATN and DN clusters in posterior cerebral cortex. Like visual areas, both DN and dATN pRFs exhibited a contralateral visual field bias, as we would predict from prior literature (31, 33, 34) (positive dATN pRFs: t(6)=26.0, p<0.001; negative DN pRFs: t(6)=4.86, p=0.003; positive DN pRFs: t(6)=5.20, p=0.002) (Fig. 1F; see Fig. S4 for additional retinotopic properties). Together, these results show that individualized DNs have voxel-wise responses that exhibit visual coding -- the necessary foundation for assessing whether the voxel-wise retinotopic code structures global interactions between the internally- and externally-oriented networks.

Visual responsiveness stratifies DN-dATN resting-state interactions

Next, we examined whether the retinotopic response properties of individual voxels (estimated during the retinotopic mapping sessions) predict their functional coupling across DN-dATN during task-absent, spontaneous activity (during the resting state scans). Specifically, we tested whether voxel-level visual response classes—non-retinotopic voxels, positively tuned pRFs (+pRFs), and negatively tuned pRFs (–pRFs)—predict the strength and direction of coupling between the DN and dATN. Such a finding would reveal a stable, voxel-level organizational principle underlying the global dissociation that is hypothesized to exist between the DN and dATN(11–13, 15, 24, 25).

Across resting state runs, global activity of the DN and dATN was independent (i.e., not correlated) after accounting for variance across all cortical networks using partial correlation(44) (mean correlation = 0. 0.031±0.13 s.d., t(6) = 0.6287, p=0.55; Fig. 2A). Crucially, however, this global relationship masked a striking stratification based on DN voxel responsiveness (Fig. 2B). Non-retinotopic DN voxels (i.e., pRF model R2<0.14) were not significantly correlated with the dATN (mean correlation = 0.014±0.135, t(6) = 0.26, p = 0.79). In contrast, DN voxels that responded positively to visual stimulation (DN positive pRFs, +pRFs) had a positive correlation with the dATN (mean correlation = 0.284±0.152, t(6) = 4.96, p = 0.0025), while DN voxels with systematic negative responses to visual stimulation (DN negative pRFs, -pRFs) were anti-correlated with the dATN (mean correlation = −0.21±0.149, t(6) = −3.75, p = 0.0094). Thus, voxel-level visual response profiles shape DN-dATN coupling during spontaneous resting-state activity. Specifically, opponency at rest between the DN and dATN appears to be driven by the subset of DN voxels with negative retinotopic tuning.

Retinotopic coding organizes spontaneous interaction between internally and externally oriented brain networks.

A. Overall activity of the dorsal attention networks (dATN-A/B) and default networks (DN-A/B) were independent at rest. One individual’s dATN and DN are shown. Time series from all voxels in the DN and dATN subnetworks were averaged together, and we calculated the partial correlation(44) of these networks while accounting for variance of the other 11 networks within the group-prior parcellation(1, 8). The dATN and DN network correlation was not significantly different from zero, suggesting the activity of these networks is largely independent. Bar plot shows the average partial correlation across participants, each data point is the average of all runs for a single participant. B. Interaction between the DN and dATN differs by visual field preference of DN voxels. Splitting the DN voxels by visual responsiveness revealed distinct patterns of resting-state correlation. DN +pRFs had a positive correlation with the dATN (mean correlation = 0.22±0.144, t(6) = 3.99, p = 0.0072), while non-retinotopic DN voxels (i.e., pRF model R2 < 0.08) were not significantly correlated with the dATN (mean correlation = 0.02±0.115, t(6) = 0.37, p = 0.72). On the other hand, DN-pRFs were anti-correlated with the dATN (mean correlation = - 0.20±0.12, t(6) = 4.22, p = 0.0055). C. Determining spatially-matched pRFs in dATNs and DNs. We assessed the influence of retinotopic coding on the interaction between internally- and externally-oriented brain areas’ spontaneous activity during resting-state fMRI, by comparing the correlation in activation between pRFs in these networks that represent similar (vs. different) regions of visual space. For each DN-pRF, we established the top 10 closest positive dATN pRFs’ centers based on their x and y position (“matched”), and correlated the DN pRF time series with the average timeseries of these matched dATN pRFs. We compared these correlation values with the correlation of the DN pRF timeseries paired with 10 randomly selected pRFs from the 33% furthest (1000 iterations per pRF). We computed this for all DN pRFs for each all runs in each subject. Inset shows the pairwise distances between all pRFs in all participants, dotted line shows the 66%ile. D. Negative (upper) and positive (lower) DN pRFs show retinotopic specific interaction with the dATN. Each set of bars depicts the average correlation of the matched and randomly paired pRFs between the DN and dATN for each participant, with each run shown as connected points. All participants showed greater correlation (either positive or negative) for matched compared with randomly paired pRFs. E. Data shown averaging runs within each participant. For both positive and negative pRFs, the correlation between matched compared with randomly matched pRFs was significantly greater (consistent with their amplitude, ps<0.01). F, G. show the correlation between positive and negative pRFs in the DN with the dATN, split across the sub-networks of the DN-A and B. Both DN-A and B show a significant impact of matching, indicating that both subnetworks interact with the dATN via a retinotopic code (ps<0.05). *: p<0.05, **: p<0.01, ***:p<0.001.

Shared visual field preferences structure functional coupling among visually-responsive DN and dATN voxels

We next asked whether functional coupling between DN and dATN voxels is modulated by the visual field preferences of retinotopic voxels — that is, whether their pRFs are centered on similar versus different visual field locations (i.e., their retinotopic alignment). We hypothesized that DN–dATN interactions would be stronger for voxel pairs with shared visual field preferences. To test this, we quantified the retinotopic alignment of voxel-wise pRFs across the DN and dATN. Because we only considered positive pRFs in the dATN, hereafter we refer to these as “dATN pRFs.”

For each pRF in the DN (+pRFs and –pRFs), we computed the Euclidean distance between its receptive-field center (x, y) and those of all dATN pRFs (Fig. 2C). We then identified the 10 closest dATN pRFs (“Matched” set) and averaged their resting-state time series. As a comparison group, we repeated this procedure using 10 randomly selected, poorly matched dATN pRFs (“Random” set; see Methods). This procedure was repeated 1000 times for each DN voxel across resting-state runs.

Matched dATN pRFs were distributed throughout the network, spanning posterior retinotopic cortex to anterior regions, including prefrontal cortex (Fig. S5). Crucially, the spatial dispersion (i.e., the distance between the pRFs across the cortex) of matched and randomly-matched pRFs did not significantly differ (matched versus random – DN-pRFs x dATN pRFs: t(6)=0.38, p=0.71; DN +pRFs x dATN pRFs: t(6)=0.67, p=0.52), ruling out spatial proximity as a confounding factor -- any difference in correlation with DN voxels should therefore be attributable to shared visual field preferences.

If DN-dATN interactions are scaffolded by a retinotopic code, spatially matched pRFs should exhibit stronger functional coupling than randomly-matched pRFs. Consistent with this prediction, after partialling out variance from DN +pRFs to isolate the specific co-fluctuation between DN - pRFs with dATN pRFs (34, 44), DN-pRFs remained anticorrelated with dATN pRFs in nearly every resting-state run (Fig. 2D). Critically, matched pRFs were significantly more anticorrelated compared with randomly-matched pRFs (linear mixed-effects model with random intercepts for subject and run: F(1,194) = 105.7, p < 0.001; coefficient = –0.023 ± 0.002 s.d.). This effect was evident when run-level correlations were averaged within each participant (7/7 participants rmatched < rrandom; t(6)=6.9, p<0.001, Fig.2E) and could not be explained by signal quality differences, as matched and random pRF sets did not differ in tSNR (t(6) = 1.01, p = 0.353). This is strong evidence that the retinotopic code underpins the opponent interaction between negative DN and positive dATN pRFs.

Voxel-wise retinotopic alignment scaffolded the interaction between positive DN and dATN pRFs as well. After controlling for DN-pRFs, DN +pRFs were positively correlated with dATN pRFs (Fig. 2D). Retinotopic alignment again strengthened the interaction: matched correlations exceeded random correlations in a mixed-effects model (F(1,194)= 406.5, p < 0.001; coefficient = +0.050 ± 0.003 s.d.) and for every individual participant (7 / 7; t(6) = 5.7, p = 0.001) (Fig. 2E).

Together, these results demonstrate that shared retinotopic alignment enhances functional coupling between DN and dATN voxels. Voxels tuned to the same region of visual space show stronger spontaneous co-fluctuation—whether positively or negatively coupled—than those with non-overlapping spatial preferences. This suggests that a shared spatial code helps organize spontaneous interactions between large-scale internal and external attention networks during rest.

Subnetwork Specificity within the DN

Recent work has suggested that the subnetworks of the DN, DN-A and DN-B, support different cognitive processes: episodic projection and theory of mind, respectively (8, 45–47). Indeed, DN-A is particularly sensitive to the process of scene construction(47), closely linked to visual imagery, suggesting that dATN pRFs would show greater interaction with visually-responsive voxels in DN-A compared with DN-B. Consistent with this hypothesis, spontaneous correlations with dATN pRFs were significantly stronger for DN-A than DN-B voxels (collapsed across positive and negative pRFs: t(6) = 4.80, p = 0.003). Crucially, however, voxel-scale retinotopic alignment structured interactions between the DN and the dATN for both DN sub-networks (Fig. 2F-G; DN-A: -pRFs t(6) = –4.56, p = 0.004; +pRFs t(6) = 5.65, p = 0.0013; DN-B: -pRFs t(6) = –7.49, p < 0.001; +pRFs t(6) = 2.92, p = 0.027). The strength of this effect did significantly differ between DN-A and DN-B (t(6) = 1.51, p = 0.182). Thus, although DN-A is more strongly globally coupled with dATN, retinotopic preferences of individual voxels structure functional interactions across both DN subnetworks equally.

In addition, using the same matched-vs-random pRF procedure, we found that resting activity between the lateral place-memory area (LPMA) and the adjacent occipital place area (OPA) was significantly more correlated for retinotopically matched than random voxel pairs (t(6)=3.01, p=0.024; SI). This shows that retinotopic scaffolding within domain-specific systems extends beyond perceptual and memory tasks (34), organizing spontaneous activity within these systems as well. These findings build on and extend our previous work (34) and others(48).

Retinotopic coding is intrinsic to internally-oriented areas

Our findings thus far highlight a role for retinotopic coding in structuring spontaneous interactions between voxels across large-scale internally- and externally-oriented brain networks. However, a key mechanistic question remains: is retinotopic coding an intrinsic property of the DN, or does it only emerge in response to bottom-up perceptual input? For instance, one possibility is that top-down DN-driven modulation of dATN activity reflects a global, non-selective suppression, with retinotopic coding in the DN arising only through bottom-up input from the dATN. Alternatively, retinotopic coding may be intrinsic to the DN itself – such that top-down signals from the DN to the dATN are retinotopically-specific, allowing the DN to shape fine-grained spatial representations within the dATN. Disambiguating these possibilities is essential for determining whether the retinotopic scaffold constitutes a general mechanism for cross-network communication, independent of task demands or cognitive domain (e.g., memory versus perception).

To address this, we again turned to participants’ resting-state fMRI data. Because participants are not engaged in any task, resting-state provides a neutral testbed for assessing spontaneous DN–dATN interactions. Moreover, experience sampling during rest has shown that attention naturally alternates between internal and external focus during rest (49, 50), making it possible to assess whether retinotopic coding emerges primarily during bottom-up (dATN-driven) or top-down (DN-driven) events.

We developed an event-triggered analytical approach to dissociate bottom-up perceptual from top-down mnemonic events. Specifically, we identified timepoints where spontaneous activity of individual DN or dATN pRFs exceeded the 99th percentile of their resting-state distribution. These high-activation time points were treated as neural “events” (Fig. 3A-C), with DN pRFs events interpreted as top-down events, and dATN pRF events as bottom-up. To test whether these events produced retinotopically-specific responses in the target network, we compared peri-event activation in the target network’s most 10 closely matched pRFs (i.e., those with similar visual field preferences) to that in the 10 worst-matched (anti-matched) pRFs (Fig. 3A). We used the worst matched pRFs for this analysis because the large number of pRFs made the random matching procedure impractical. All analyses were conducted separately for positive and negative pRF populations within the DN. This design enabled us to test two competing predictions: if retinotopy in the DN only reflects bottom-up input, we should only see retinotopically specific coupling during bottom-up (dATN-driven) events. If, however, the DN possesses an intrinsic retinotopic scaffold used for top-down communication, we should observe retinotopic specificity during both bottom-up and top-down events. A windowed correlation analysis produced similar results (Fig. S6).

Top-down vs. bottom-up neural events detected in spontaneous resting-state dynamics show evidence for retinotopically-specific suppression.

A. Event detection and analysis procedure and example events from a single resting-state run. To detect events, we extracted the time series from each pRF in the source regions (top-down: positive or negative pRFs in the DN; bottom-up: positive pRFs in the dATN) and isolated time points where the z-scored time series exceeded 2.4 s.d. (99th percentile). We then examined the activity of 10 best matched and 10 worst matched (anti-matched) pRFs from the target region in this peri-event time frame (6 TRs (8 s) before and after the event). B. Example top-down event in the representative time series shows elevated source pRF activity (DN-pRF) and corresponding target area activity (dATN +pRF). C. Same as B, but for bottom-up activity with the event detected in a dATN +pRF. D, E. Evidence for retinotopically-specific suppression of ongoing activity during events top-down events detected in DN-pRFs (D) and bottom-up events detected in dATN +pRFs (E). D. Time series show the peri-event activity of matched and anti-matched dATN +pRFs averaged across all participants for events detected in DN-pRFs (N=78,978 events). E. Time series shows the peri-event activity in DN-pRFs during events detected in dATN +pRFs (N=541,546). F. For DN-pRFs paired with dATN +pRFs, target area shows retinotopically-specific suppression of activity for both top-down and bottom-up events. Bars show the average activation at event time of the target areas’ matched and anti-matched pRFs for each participant. Activity was lower in matched compared with anti-matched pRFs for top-down and bottom-up events (ts>3.8, ps<0.009). G, H. Evidence for retinotopically-specific excitation of ongoing activity during events top-down events detected in DN +pRFs (G) and bottom-up events detected in dATN +pRFs (H). G. Grand average time series for matched and anti-matched dATN +pRFs during events detected in DN +pRFs for all participants. H. Grand average time series for matched and anti-matched DN +pRFs during events detected in dATN +pRFs. I. For DN +pRFs paired with dATN +pRFs, target area shows retinotopically-specific excitation for both top-down and bottom-up events. Bars show the average activation at event time of the target areas’ matched and anti-matched pRFs for each participant. Activity was higher in matched compared with anti-matched pRFs for top-down and bottom-up events (ts>8.75, ps<0.001). Interestingly, the activity of anti-matched pRFs was greater during top-down compared with bottom-up events (t(6)=4.92, p=0.0026), suggesting that top-down input from DN +pRFs to the dATN +pRFs may be more global and less retinotopically-targeted compared with bottom-up signaling between these networks. For time series plots, solid bars along the x-axis indicate periods with a significant difference between matched and anti-matched pRF activity within the target area. *: p<0.05, **: p<0.01, ***:p<0.001. -DN: Negative DN pRFs; +DN: positive DN pRFs; +dATN: Positive dATN pRFs.

Across all participants, we detected 78,978 top-down events in DN-pRFs, 52,595 top-down events in DN +pRFs, and 541,546 bottom-up events in dATN pRFs. Bottom-up events were more frequent than either top-down type (dATN vs. negative DN: t(6)=15.19, p<0.001; dATN vs. positive DN: t(6)=25.63, p<0.001), but there was no difference between top-down event types (t(6)=2.03, p=0.18). On average, participants exhibited 2082±535 (mean±SD) negative DN, 1799±400 positive DN, and 3102±602 dATN events per run. Each pRF showed 0–4 events per run (Fig. S7). More than 90% of timepoints involved <10 simultaneous events, suggesting that individual pRF-level correlation could be isolated from global fluctuations in regional activity, making this approach suitable for evaluating distinct interactions at the individual voxel level.

We found a robust effect of retinotopic alignment (i.e., matched versus anti-matched) on pRF activation during both bottom-up and top-down interactions (Fig. 3D-I). For DN-pRFs, a linear mixed-effects model (factors: top-down/bottom-up and retinotopic alignment; subject as random effect) revealed a main effect of match status (F(1,24)=15.3, p=0.0006) and no interaction between match status and direction (F(1,24)=0.0748, p=0.40). During top-down events in DN - pRFs, matched dATN pRFs showed lower activity than anti-matched (t(6)=5.46, p=0.001) and were significantly deactivated relative to the pre-event baseline. The same was true of DN-pRFs during bottom-up events in dATN +pRFs: we observed lower activation for matched vs. anti-matched pRFs (t(6)=3.80, p=0.009). These results are consistent with a retinotopically-specific interaction between dATN +pRFs and DN-pRFs during both top-down and bottom-up events detected in spontaneous activity during resting-state fMRI.

The interaction between DN +pRFs and dATN pRFs also showed evidence of retinotopic coding during both bottom-up and top-down events, but these effects were modulated by a direction × match interaction (F(1,24)=12.84, p=0.002). Both top-down (t(6) = 8.76, p < 0.001) and bottom-up (t(6) = 9.14, p < 0.001) events evoked strong activation in the target network (dATN or DN, respectively). However, anti-matched pRFs were more active during top-down compared to bottom-up events (t(6) = 4.92, p = 0.0026), suggesting that top-down signals from positive DN pRFs to dATN may be less retinotopically specific than bottom-up signals. These results align with prior findings showing decreased spatial precision in visual representations during memory recall compared to perception(51, 52).

Taken together, these results support the view that retinotopic coding serves as a general scaffold for DN-dATN interactions. This alignment shapes spontaneous coupling during both periods of DN-driven (top-down) and dATN-driven (bottom-up) activity, even in the absence of external tasks. Because the effect persists without sensory input, it likely reflects an intrinsic feature of the DN activity rather than a transient result of bottom-up drive.

Discussion

Despite extensive evidence that internally- and externally-oriented brain networks – including the default network (DN) and dorsal attention network (dATN) – are critical for human cognition(1–8, 11–13), the coding principles that enable these distributed systems to interact have remained unclear(30, 53–57). Here, we show that DN–dATN coupling is retinotopically grounded at the voxel scale: the visual field biases of individual voxels scaffold bidirectional interactions between these networks, even during spontaneous dynamics in the absence of visual task demands. Analyses of transient elevated activity in either the DN or dATN confirm that this retinotopic opponency shapes both bottom-up signals (from the dATN to the DN) and top-down signals (from the DN to the dATN), indicating that the retinotopic code is intrinsic to both internal and external networks. These findings offer a multi-scale account of network communication in which macro-scale dynamics are underpinned by latent, fine-grained voxel-scale biases. Rather than functionally dissociated, our data show that the DN and dATN are dynamically coordinated systems that are poised to integrate internal and external information through a shared spatial scaffold.

Resolving the coding principles of the DN is critical given its central roles in spatial and episodic memory, social cognition, and executive function(13, 58–63). Classic views of systems-level brain organization posit regions at the apex of the cortical hierarchy, like the DN, likely have abstract, amodal neural codes, in contrast to sensory codes–like retinotopy–that manifest in earlier areas of the cortical hierarchy (2, 29, 30). Our findings and others challenge this view. Instead of shedding the retinotopic code at the cortical apex, the DN retains a retinotopic code(31, 33) and functionally deploys this code to structure interactions with the externally-oriented dATN. Thus, a code once thought to be exclusive to early visual cortex and visually-oriented tasks plays a widespread role in shaping brain activity across the cortical hierarchy—even at rest.

Establishing this voxel-level structure within the DN and dATN reframes a long-standing debate that has cast these networks as globally competitive or functionally independent(11, 23, 27, 28). These views have stemmed from macro-scale observations of reciprocal activation patterns during external versus internal task demands(6, 10, 13, 14, 16–23), and resting-state analyses that reveal zero or negative correlations between their mean time series(11, 12, 24, 25). However, by focusing on global, network-level activity across voxels, these approaches have obscured a latent, voxel-level structure to DN-dATN functional coupling. Specifically, we find that DN–dATN coupling is not globally antagonistic. Instead, two subpopulations of voxels within the DN contribute to a retinotopically-grounded cross-network interaction that is stratified by their voxel-level visual selectivity. One population of DN voxels is positively tuned to visual stimulation, and positively covaries with dATN voxels that share their visual field preference. In contrast, DN voxels that are negatively tuned to visual stimulation negatively covary with retinotopically matched dATN voxels as a function visual field preference. When these voxel-level tuning preferences are not taken into account, cross-network activity “averages out” into a global opponency or independence(11, 12), reinforcing the misconception that the networks support mutually exclusive cognitive modes. By revealing a shared retinotopic scaffold, our findings reconcile why DN and dATN can appear oppositional in coarse analyses yet remain functionally integrated, offering a new framework for understanding how internal and external attention are coordinated in the human brain.

Importantly, there are key differences in the nature of retinotopic coding between the DN and visual areas, including the dATN. First, retinotopy in the DN is predominantly negative in valence—DN voxels more often show spatially-specific suppression in response to visual stimulation(31, 33, 34), unlike the positive spatially-specifical response that is typically observed in visual areas. Second, although many DN voxels exhibit reliable retinotopic coding, the DN lacks the map-like retinotopic organization seen in early visual cortex. In visual areas, retinotopic coding is not only the foundational code, but also the dominant topographic organizing principle across the cortical surface, where voxels that are proximal on the cortical surface tend to share visual selectivity for nearby regions of visual space(64). However, the presence of a coding scheme does not necessarily require a map-like organization on the cortical surface. For example, rodent and bat head direction cells are not laid out in a literal ring, but the population code of these neurons forms a ring manifold(65, 66). Similarly, in humans, retrosplenial cortex shows head-direction coding during navigation tasks despite lacking topographic organization of head-direction responses(67). Analogously, the DN may implement a retinotopic code that is distributed across the network in a format that does not spatially mirror the retina. That is, the DN may use a retinotopic code without a retinotopic map – the representational structure can be preserved even in the absence of spatially ordered cortical layouts.

What function might the bivalent (positive and negative) voxel-scale retinotopic alignment across these large-scale networks serve? Our previous work has identified a role for retinotopic-alignment in structuring the functional coupling of scene-selective mnemonic and perceptual brain areas(34, 48) located near the dATN and DN, respectively (36, 68). Near to the DN, scene-selective memory areas exhibit a negative retinotopic code, while functionally-paired scene-selective perceptual regions exhibit a positive retinotopic code(34). This bivalent retinotopic code structures the mutual activity of these mnemonic and perceptual areas: scene-recall activates scene-memory voxels, it but suppresses voxels in scene-perception areas that are tuned to matched visual locations, a pattern that reverses during scene perception(34). The current work significantly extends this observation beyond perceptual and mnemonic task-states -- identifying widespread retinotopic alignment across large-scale networks with presumed different orientations (internal vs. external) even during spontaneous activity at rest, in the absence of tasks.

As to the functional role of the bivalent retinotopic alignment across internal and externally-oriented large-scale networks, we propose two complementary possibilities. One possibility is that the bivalent (positive and negative) subpopulations of DN pRFs may serve as a neural mechanism for maintaining separate channels for perceptual and mnemonic information(54), preventing mnemonic signals from being interpreted as or interfering with sensory signals. A second possibility is that bivalent coding across the DN and dATN could be a mechanism for signaling mnemonic predictions via top-down inhibitory processes(34, 69). Specifically, when the DN generates predictions about incoming sensory information based on memory, the inverted retinotopic code could act as a suppressive signal that sharpens tuning in perceptual areas and enhances signal-to-noise for expected stimuli(70, 71). These mechanisms may work in concert: the inverted code simultaneously maintains separate channels for memory/perception within the DN, while enabling precise top-down predictive signaling across DN-dATN. Future work that combines high-resolution recordings with behavior will be essential for testing how these mechanisms jointly support constructive operations such as memory-based prediction during visually guided navigation.

Retinotopic alignment between the DN and dATN was clear in both subnetworks of the DN, DN-A and DN-B(3, 8, 20). This is surprising given their distinct functional profiles: DN-A has been more directly linked to visually grounded memory processes, such as episodic projection and scene construction(8, 20, 47), and spatially overlaps with regions previously shown to exhibit retinotopy and engage in scene memory and perception(68). In contrast, DN-B is more often associated with processes that are not obviously tied to vision: theory of mind and affective tasks, such as interpreting false beliefs or representing physical and emotional pain (8, 20). Why might the DN-B show retinotopically-grounded coupling with perceptual networks? Retinotopic coding might still be a useful format for transforming abstract social representations into sensory-grounded signals to inform perceptual predictions (e.g., anticipating changes in facial expression) or behavioral decisions (e.g., making eye movements). Future work might consider direct manipulation of social perceptual processes to assess the impact of social tasks on retinotopic coding between DN-B and dATN.

Moments of high spontaneous activity originating in the DN co-occurred with suppression of retinotopically-matched dATN pRFs, suggesting that retinotopy functions as a native coding format in the DN for targeted modulation of perceptual cortex. This result aligns with other recent studies suggesting an important role of the DN in shaping visual processing. For example, ongoing pre-stimulus DN activity has been shown to influence sensitivity during near-threshold visual object recognition(72). Other work has shown that subareas of the DN encode kinematic of spatial exploration in rodents(73), and represent the visuospatial context that is associated in memory with a perceived scene(34, 36, 68, 74, 75). Relatedly, DN activity reflects semantic-level attentional priorities during visual search(76). Taken together, these findings show that, rather than being disengaged during visual tasks, the DN actively shapes responses in perceptually-oriented cortex in a spatially-precise manner. In this sense, retinotopic coding is part of the DN’s “native language”. Taken together, these findings prompt a reevaluation of the DN’s role in perceptual processing, and the extent to which roles in “internally- and externally-oriented attention” adequately captures the DN and dATN’s role in cognition.

Is the DN a “visual” network? One possible interpretation of retinotopic coding in the DN is that the DN directly and obligatorily represents retinal position in addition to more abstract features(38, 64). Under this hypothesis, the retinotopic code may indicate that the information represented in the DN is, in part, sensory. Alternatively, the apparent visual coding may simply represent an underlying connectivity structure(77), which enables associative and semantic information that the DN directly represents(30, 78, 79) to be effectively transferred to sensory areas of the brain. Under this alternative hypothesis, the information represented in the DN is not itself sensory in nature; the retinotopic code in the DN simply serves as a “highway” between the abstract representations in the DN and sensory representations in the dATN. Notably, in either case the retinotopic scaffolding represents a latent structure linking high-level and low-level areas through consistent, spatially-organized interactions(77). This provides more evidence that low-level sensory codes are distributed across the brain, joining recent work showing latent somatotopic codes in regions of the brain that are not typically associated with somatosensation, including visual cortex(80). The wide and overlapping distribution of sensory codes across the brain may allow for multiplexing of complex information to structure cross-network interactions(35, 80). Further studies investigating this framework with diverse tasks and stimuli could reshape our understanding of how the brain processes and integrates information across different levels of cognitive complexity.

In summary, our results show that a voxel-level retinotopic code organizes the spontaneous interactions between large-scale internally- and externally-oriented networks in the human brain – the DN and dATN. Even when global activity appears uncorrelated across DN-dATN, voxel-level visual-field tuning reveals a tightly interlocked, bidirectional functional dynamic that persists in the absence of external stimulation. This latent scaffold reconciles the apparent macroscale independence of internally- and externally-oriented systems with their evident functional need to cooperate, revealing that the DN is not divorced from perceptual networks, but modulates perceptual voxels through a common visuospatial reference frame. Taken together, by grounding large-scale network interactions in a shared, bivalent spatial code, retinotopy may offer a simple organizing principle for the brain’s coordination of thought and perception.

Materials and Methods

Dataset and Participants

The data analyzed here are part of the Natural Scenes Dataset (NSD), a large-scale precision 7T fMRI dataset consisting of high-resolution anatomical scans, retinotopic mapping, functional localizers for visual areas, and both task and resting-state fMRI data for eight adult participants. A full description of the dataset can be found in the original manuscript (37). Here, we detail the data, processing, and analysis steps relevant to the present work.

Participants (N = 8; 2 male, 6 female; ages 19–32) were scanned at the University of Minnesota over the course of 33-43 sessions. All participants had normal or correct to normal vision and no known neurological impairments. Informed consent was collected from all participants, and the study was approved by the University of Minnesota institutional review board. One subject (subj03) was excluded from resting-state analyses for having insufficient resting-state runs that passed our quality metrics (see below).

For the present study, we focused on two components of the NSD protocol: retinotopic mapping and resting-state fMRI. All participants completed three runs of a sweeping bar retinotopic mapping task (5 mins each). Retinotopic mapping was conducted during Session 1, along with the functional localizer scans. This task is referred to as “RETBAR” in NSD documentation.

All participants completed at least 14 runs of a resting-state fixation task (range: 14-34 runs; 5 mins each). Resting-state data runs were included in all NSD scan sessions after session 20 (i.e., session 21, 22, etc.). Each of these sessions included two 5-minute resting-state runs -- one at the start and one at the end of the session -- yielding 14 to 34 total resting-state runs per participant. During the second resting-state run, participants performed a deep breath in the first 5 seconds of scanning. To account for this, and to ensure the two resting-state runs were equated in length, we excluded the first 30 seconds of the resting-state data from all runs. One participant (subj03) was excluded from resting-state analyses due to having an insufficient number of usable runs after quality control. All remaining participants had at least 8 high-quality resting-state runs (mean: 14 ± 6.13 runs).

MRI acquisition and processing

For this study we made use of the following data from the NSD: anatomical (T1 and FreeSurfer segmentation/reconstruction(81, 82)), functional regions of interest (ROIs), and minimally preprocessed retinotopy and resting-state time series. All analyses except network identification were conducted in original subject volume space, and data were projected on the surface for visualization purposes only. Network identification was performed on the fsaverage6 surface, then projected into subject specific volume space for quantification.

Anatomical data

Anatomical data was collected using a 3T Siemens Prisma scanner and a 32-channel head coil. We used the anatomical data provided at 0.8mm resolution as well as the registered output from FreeSurfer recon-all, aligned to the 1.8mm functional data. For visualization purposes, we projected statistical and retinotopy data to the cortical surface using SUMA (83) from the AFNI software package(84).

Functional MRI data acquisition and processing

All analyses were conducted on 1.8mm isotropic resolution minimally processed runs of a sweeping-bar retinotopy task and resting-state fixation provided in the NSD.

Quality assessment

To ensure only high-quality resting-state data were included, we trimmed the first 25 TRs (40s) from each run (74), which left 4.25 min of resting-state data per run. We then used AFNI’s quality control assessment tool (APQC(85)) on the raw trimmed resting-state time series to assess the degree of motion in the resting-state scans. We excluded runs with greater than 1.8mm maximum displacement per run or 0.12mm framewise displacement from analysis, and assessed runs with greater than 1mm displacement or 0.10mm framewise displacement on a case-by-case basis(8). After exclusions, one participant (subj03) had only a single run that survived our criteria (4.25 minutes of data)), so we excluded them from our analyses. The remaining 7 participants had at least 8 resting-state runs (> 34 minutes of data; mean number of runs: 14±6.13 (sd), range: 8-24 runs).

ICA denoising

To further denoise the retinotopy and remaining resting-state data, we used manual ICA classification of signal and noise on the minimally-preprocessed time series(36, 40, 41). We used manual classification because automated tools perform poorly on the high spatial and temporal resolution data of the NSD(86). For each retinotopy and resting state run, we decomposed the data into independent spatial components and their associated temporal signals using ICA (FSL’s melodic(87, 88)). We then manually classified each component as signal or noise using the criteria established in(41). Noise signals were projected out of the data using fsl_regfilt(40).

We did not perform global signal regression(26). Data were normalized to percent signal change. A 2.5mm FWHM smooth was applied on the surface to resting-state fixation data used in functional connectivity network identification. The analysis of voxel-scale interactions performed on unsmoothed data.

Data analysis – Identifying Resting-State Networks

Individual-specific cerebral network estimation

In each participant, we identified a set of 15 distributed networks on the cerebral cortical surface using a multi-session hierarchical Bayesian model (MS-HBM) functional connectivity approach (8, 39). Briefly, the MS-HBM approach calculates a connectivity profile for every vertex on the cortical surface derived from the top 10% of correlations from that vertex to 1,175 ROIs uniformly disturbed across the cortical surface(1). We applied an expectation-maximization algorithm, initialized with a 15-network group-level spatial prior created by applying previous methods of network identification(1) on a portion of the HCP S900 data, to all the connectivity profiles from all participants. This algorithm then estimated each network’s group-level connectivity profile, along with parameters for subject-, session/run-, and region-level variability, which together were used to generate individual-specific, labeled parcellations.

The MS-HBM approach is well-suited to derive parcellations for the participants in the NSD, where resting-state data was collected across multiple days, because this method uses repeated resting-state runs for each participant to estimate within-participant variance. This leverages a strength of the NSD (the relatively large number of resting-state runs for each individual) and can account for differences in connectivity both within an individual (likely the result of confounding variables such as scanner variability, time of day, etc.), and between participants (reflecting potentially meaningful individual differences). Moreover, the MS-HBS incorporates a group spatial prior, further stabilizing the resulting network parcellations for each participant. This also ensures that all networks will be identified in all participants, while allowing for idiosyncratic topographic differences.

Data analysis – Retinotopic Mapping

Population receptive field (pRF) Modelling

The NSD retinotopy stimuli consist of a mosaic of faces, houses, and objects superimposed on pink noise that are revealed through a continuously drifting aperture. For our analysis, we considered only the bar stimulus time series, which is consistent with other studies investigating negative pRFs in high-level cortical areas (31, 33, 34). We did not consider the wedge/ring stimulus for any analyses.

After denoising the retinotopy data, we averaged the three retinotopy runs with the bar aperture together to form the final retinotopy time series. We performed population receptive field modeling using AFNI following the procedure described in (43). First, because the pRF stimulus in the NSD is continuous, we resampled the stimulus time series to the fMRI temporal resolution (TR = 1.333s). Next, we implemented AFNI’s pRF mapping procedure (3dNLfim). Given the position of the stimulus in the visual field at every time point, the model estimates the pRF parameters that yield the best fit to the data: pRF amplitude (positive, negative), pRF center location (x, y) and size (diameter of the pRF). Both Simplex and Powell optimization algorithms are used simultaneously to find the best time series/parameter sets (amplitude, x, y, size) by minimizing the least-squares error of the predicted time series with the acquired time series for each voxel. Relevant to the present work, the amplitude measure refers to the signed (positive or negative) degree of linear scaling applied to the pRF model, which reflects the sign of the neural response to visual stimulation of its receptive field.

Identifying Retinotopic Voxels – Empirically-derived R² thresholds

We developed a rigorous, data-driven approach to determine whether our pRF model fits truly reflect visual coding rather than physiological noise. To do this, we leveraged the resting-state fMRI data – where there is no dynamic visual stimulation -- to create a control (null) distribution for retinotopic model fits and establish a “noise floor” for each network (i.e., the DN and dATN). Then, we compared task-derived model fits to this rest-based null distribution to establish an empirical R² threshold for pRFs in each network.

Specifically, we fit the pRF model to the average timeseries across all resting-state runs for each participant. We then compared the resulting voxel-level R² distribution to that from the pRF mapping task data, pooled across participants. To test whether the model fits were significantly higher for the task vs. null distribution, we performed a one-tailed Kolmogorov-Smirnov test for each network on the respective cumulative distribution functions (H0: CDFrest(R2) ≤ CDFtask(R2), Ha: CDFrest(R2) > CDFtask(R2)). These analyses were performed separately for the dATN and positive and negatively responsive voxels within the DN.

We defined a voxel as exhibiting a retinotopic code (hereafter referred to as a “pRF”) if its model fit exceeded 14% variance explained. This threshold was based on the 95th percentile of the null R² distributions, which yielded noise floors of R² = 0.124 (dATN), R² = 0.132 (DN-positive), and R² = 0.137 (DN-negative). We therefore set our global threshold for visually responsive voxels across all networks at R² > 0.14, as this value exceeded the noise floor calculated in each network separately.

Quantifying visual field coverage

Visual-field coverage (VFC) plots represent the sensitivity of an ROI across the visual field. We followed the procedure in Steel et al., 2024(34), which we have reproduced here. Individual participant VFC plots were first derived based pRF center, as well as pRF size and R2. These plots combine the best Gaussian receptive field model for each suprathreshold voxel within each ROI by calculating the max operator at each point in the visual field, defined as the maximum value from all pRFs within the ROI. The resulting coverage plot thus represents the maximum envelope of sensitivity across the visual field. Individual participant VFC plots were averaged across participants to create group-level coverage plots.

To compute the elevation biases, we calculated the mean pRF value (defined as the mean of the average max operator values in a specific portion of the visual-field coverage plot) in the contralateral upper visual field (UVF) and contralateral lower visual field (LVF) and computed the difference (UVF–LVF) for each participant, ROI and amplitude (+/−) separately. A positive value thus represents an upper visual-field bias, whereas a negative value represents a lower visual-field bias.

Reliability of pRF amplitude and center position estimates

We tested whether key pRF features of interest – amplitude (i.e., positive vs. negative) and center position (i.e., x,y) – were reliable across retinotopy runs. To assess the reliability of pRF amplitude, we iteratively compared the amplitude of significant pRFs from individual runs of pRF data. Specifically, for each retinotopy run, we fit the pRF model and then binarized any significant voxels (R² > 0.14) by amplitude. When testing positive-pRF reliability (in dATN and DN), positive voxels = 1; all others = 0. For negative-pRF reliability (in DN), significant negative voxels = 1. We then computed a three-run Dice coefficient: . We compared this value against 5000 bootstrapped Dice scores derived from the same number of voxels randomly sampled from all voxels, both significant and non-significant, in the ROI. For each participant, this resulted in 1 “observed” dice coefficient along with 5000 bootstrapped values representing the distribution of dice coefficients expected by chance. The observed coefficient was z-scored against this null distribution for each participant.

We also evaluated center position (x,y) reliability. For every significant voxel, we calculated the Euclidean distance between its center estimates across the three run pairs (i.e. 1-2, 2-3, 1-3). As with amplitude, we generated a null distribution by repeating this computation on 5,000 random voxel sets (significant + non-significant). To establish significance of the true estimate, the observed estimate was z-scored against this null distribution for each participant.

Data Analysis – Retinotopically-Grounded Resting-state Analyses

Evaluating global functional coupling between cortical networks at rest

To assess global functional coupling between the default network (DN) and the dorsal attention network (dATN) at rest, we used a partial correlation analysis that accounted for variation associated with other cortical networks (44). For this and all subsequent analyses, each individual resting state run was z-scored in time. Prior to calculating the partial correlation between them, we first averaged the time series within each DN and dATN subnetwork.

Evaluating the Contributions of Voxel-Level Visual Response Valence to Cross-Network Coupling

To test whether DN–dATN coupling is modulated by the visual responsiveness (i.e., response valence) of DN voxels, we repeated this analysis separately using only: (i) sub-threshold/non-retinotopic voxels, (ii) positive pRFs, or (iii) negative pRFs from the DN -- while still partialing out the remaining networks. This allowed us to assess whether specific classes of visually responsive voxels in the DN differentially contribute to functional coupling with positive pRFs in the dATN.

Evaluating Voxel-Level Retinotopic Alignment in Cross-Network Coupling

To test whether functional coupling is shaped by the shared visual field preferences of DN-dATN voxels (i.e., retinotopic alignment), we compared resting-state correlations between pRF pairs across the DN and dATN that had matched (vs. unmatched) pRF centers.

To do this, for each pRF in the DN (positive or negative), we computed the Euclidean distance between its receptive-field center (x, y) and those of all dATN +pRFs. We focused on center position instead of pRF size because 1) size estimates are relatively unstable compared with center position(89), and 2) we lacked a clear prediction about how pRF size would be related between the two networks. For each DN voxel, we identified the 10 nearest dATN +pRFs based on center position—the “matched” set—and averaged their resting-state time series. As a null comparison, we defined a “random” set by sampling with replacement 10 dATN +pRFs from the most distant third of the match distribution (i.e., (the 1/3 furthest pRF pairs). We considered the bottom 33% because the overwhelming majority of pRFs constituted relatively equal “good” matches. This random sampling was repeated 1,000 times per DN voxel, generating a distribution of 1,000 random correlation values, which we then averaged to yield a single “random” correlation for each DN source pRF.

We calculated the correlation between each DN pRF’s resting-state time series and that of its 10 best-matched dATN pRFs (“matched” correlation). We averaged across the 10 best matches to (1) reflect the likely distributed input from multiple dATN sources and (2) improve signal-to-noise. To isolate pRF-specific effects, we partialed out the mean time series of the complementary DN pRF population from the signals of interest. For example, when analyzing correlations between – DN and +dATN, we treated +DN activity as a nuisance regressor (and vice versa)(34, 44). This provided a network-level analog to the whole-brain partial correlation described above (see Section: Correlation among cortical networks), allowing us to isolate the unique contribution of the pRFs of interest from overall activity in the region as well as to control for global fluctuations that cause widespread positive correlations like motion and attentional state(34, 44). All correlation values were Fisher-transformed, and we took the median value across voxels within each network to yield a single value per run. This procedure was repeated for all runs and all participants.

We then statistically compared Fisher-transformed correlations for matched vs. random pRF pairs using a linear mixed effects model, with match status (matched vs. random) as a fixed effect and subject and run as random effects. This model accounted for both between- and within-subject variation across session. For follow-up tests, we compared the average matched and random values across participants using a paired t-test.

Resting-state event detection analyses

To dissociate the influence of top-down versus bottom-up drive on spontaneous interactions between the DN and dATN, we developed an event-triggered analysis of resting-state fMRI data. Because participants were not engaged in any task, resting-state data provide a neutral testbed for evaluating the directionality and spatial specificity of DN–dATN interactions. Building on prior findings that attention during rest naturally alternates between internally and externally oriented states(23, 49, 50) we reasoned that transient “neural events” in these networks could reveal spontaneous top-down and bottom-up communication.

We defined top-down events as transient high-activation periods in DN pRFs, and bottom-up events as high-activation periods in positive-polarity dATN pRFs(90, 91). Event detection was performed independently for each source pRF by taking each pRF’s z-scored resting-state time series and identifying TRs with unusually high activity. Specifically, we considered any time point with a z-value greater than 2.4 (i.e., 99.18 percentile) as a neural event. Results were comparable with a range of thresholds (2.1 < z < 2.9).

At each detected event, we extracted the resting-state time series of the 10 most retinotopically aligned pRFs in the target network—i.e., the 10 with the smallest Euclidean distance in receptive-field center (x, y)—over a 13-TR peri-event window (–6 to +6 TRs). These time series were normalized to the mean of the first four pre-event TRs to allow comparisons across events and conditions. To evaluate the spatial specificity of these interactions, we compared responses in the matched pRFs to those in an “anti-matched” pRFs set: the 10 most distant pRFs in the target network in terms of visual field preference (i.e., pRF center positions). Unlike prior analyses that used random sampling to define unmatched control pRFs, here we used a deterministic anti-match procedure to avoid the computational burden of repeated resampling, given the large number of events.

This procedure was repeated across all relevant source–target network pairs: top-down (DN as source, dATN as target), bottom-up (dATN as source, DN as target), and for both positive and negative DN pRFs. For each event type and match status, we computed the average response in the target pRFs at time zero (event peak). These values were entered into a linear mixed effects model with fixed effects of match status (matched vs. anti-matched) and event direction (top-down vs. bottom-up), and with subject as a random effect. This allowed us to test whether retinotopic alignment modulates DN-dATN interactions during spontaneous events and whether such modulation differs for top-down vs. bottom-up events.

Statistical tests

Statistical analyses were implemented in MATLAB (Mathworks, Inc). Given the relatively small number of participants and large number of runs per subject, we used linear mixed effects models to account for both within- and between-subject variability, as described in the relevant sections above. For follow-up tests, we used paired-sample t-tests and corrected for multiple comparisons where appropriate. Corrected p-values are reported where appropriate.

eLife Assessment

This study addresses an important question about how large-scale brain networks interact, and specifically how the default mode network exchanges information with the sensory cortex. The analyses are sophisticated, but at present provide incomplete evidence for the claims made in the paper.

Reviewer #1 (Public review):

Summary:

This paper leverages 7T fMRI data from the Natural Scenes Dataset to investigate whether retinotopic coding, the position-selective organization of visual response structures, spontaneous resting-state interactions between the Default Network (DN) and the Dorsal Attention Network (dATN). Using individualized network parcellations and population receptive field (pRF) modeling, the authors show that DN voxels can be split into two subpopulations based on their response to visual stimulation: those with position-specific positive BOLD responses (+pRFs) and those with position-specific negative BOLD responses (-pRFs). Critically, these subpopulations relate differently to the dATN during rest: -pRFs are anticorrelated with the dATN, +pRFs are positively correlated, and non-retinotopic DN voxels show no coupling. The anticorrelation (and positive correlation) is enhanced when DN and dATN voxels share visual field preferences. An event-triggered analysis suggests that retinotopic coding shapes both "top-down" (DN-initiated) and "bottom-up" (dATN-initiated) spontaneous activity transients, supporting the claim that the retinotopic scaffold is intrinsic to the DN. These findings challenge the prevailing view of global DN-dATN antagonism and suggest retinotopic coding as an organizing principle for cross-network communication.

Strengths:

The central finding that what looks like network-level independence between DN and dATN decomposes into structured, bivalent interactions organized by voxel-level visual field preferences is a compelling demonstration that macro-scale network descriptions can hide meaningful substructure. The logic of the analysis is clean: pRF properties are estimated from retinotopic mapping data and then used to predict resting-state coupling in completely independent scanning sessions. This cross-session, cross-modality design rules out many circularity concerns.

The use of individualized multi-session hierarchical Bayesian parcellation (Kong et al.) to define DN and dATN boundaries within each subject is the right methodological choice for this question. Network boundaries in posterior cortex, where DN and dATN interdigitate most closely, vary considerably across individuals, and group-average approaches would introduce exactly the kind of misassignment that would most confound the result.

The matched-vs-random pRF analysis is well-controlled. The authors demonstrate that cortical distance between matched and randomly-matched dATN pRFs does not differ, effectively ruling out spatial proximity on the cortical surface as a confound. tSNR controls further show that signal quality differences do not drive the effect.

The event-triggered analysis (Figure 3) is creative and adds genuine value. Showing that retinotopically-specific coupling persists during DN-initiated activity transients, not only dATN-initiated ones, is the key piece of evidence for the claim that the code is intrinsic to the DN rather than passively inherited through bottom-up visual drive.

The result is observed consistently across all individual participants, which provides strong evidence for the robustness of the qualitative pattern despite the small sample size inherent to densely-sampled designs.

Weaknesses

(1) The nature of negative pRFs requires more scrutiny

The entire interpretive framework depends on treating negative pRFs in the DN as genuine position-selective neural responses (suppression). However, negative BOLD signals are well known to arise from non-neural sources, specifically, vascular stealing (where activation in nearby tissue diverts blood from adjacent voxels) and macrovascular draining vein effects that produce spatially displaced signal inversions. These concerns are amplified at 7T, where T2*-weighted GE-EPI carries substantial macrovascular weighting. The DN and dATN interdigitate extensively in the posterior cortex, often within millimeters. A negative pRF in a DN voxel adjacent to a positive dATN voxel could, in principle, reflect the hemodynamic shadow of its neighbor rather than an independent neural response.

The spatial dispersion control (matched vs. random pRFs have similar cortical distribution) is valuable but addresses long-range confounds, not *local* hemodynamic crosstalk. The reliability of sign and center position across runs is reassuring but does not exclude a vascular origin, as vascular architecture is itself stable across sessions. I would encourage the authors to test whether the matched-vs-random effect survives exclusion of voxels near large pial vessels (identifiable from T2* contrast or the venograms available in the NSD). These analyses would not be dispositive, but they would meaningfully strengthen the neural interpretation.

(2) Amount of retinotopic mapping data and choice of pRF pipeline

The NSD includes 6 runs of retinotopic mapping (~5 minutes each; 3 bar-aperture, 3 wedge/ring). The authors use only the 3 bar-aperture runs (~15 minutes total per subject) and fit their own pRFs using AFNI's 3dNLfim procedure, rather than using the pRF estimates provided as part of the NSD release (which were fitted using the analyzePRF toolbox with all 6 runs).

Fifteen minutes of bar data is quite limited for reliable voxel-wise pRF estimation, especially in regions far from the early visual cortex, where signal-to-noise is inherently lower. Standard recommendations for robust pRF mapping in higher-order regions generally suggest substantially more data. The variance-explained threshold is close to the noise floor by design, meaning that a non-trivial number of the "retinotopic" DN voxels may be poorly estimated. Given that the core analyses depend on both the sign and the center position of these pRFs, the limited data is a significant concern.

The authors do not explain why they chose to re-fit pRFs rather than use the NSD-provided estimates. If the motivation was methodological (e.g., the NSD pRF pipeline does not readily yield signed amplitude, or the bar-only fits were judged more appropriate for detecting negative responses), this should be made explicit. If the NSD-provided pRFs can reproduce the key findings, this would substantially increase confidence in the results. If they cannot, that divergence itself would be important to understand. I would ask the authors to address this choice and, if feasible, to report whether the core results replicate using the NSD-provided pRF estimates and/or whether using all 6 runs of retinotopy data changes the findings.

(3) pRF model adequacy for the Default Network

The isotropic Gaussian pRF model was developed for and validated in early and mid-level visual cortex, where it captures the dominant spatial selectivity of neuronal populations. In DN voxels where the model explains comparatively little variance, it is less clear that the model is capturing the right quantity. Specifically, the negative pRFs could conceivably be described by a model with a dominant suppressive surround (e.g., a difference-of-Gaussians model), in which what appears as a "negative pRF" in the standard model is actually the surround component of a center-surround mechanism whose center is poorly resolved. This distinction matters: a genuine inverted code (negative center response) implies a qualitatively different computation than inherited surround suppression from nearby visual cortex.

The authors should consider discussing why the standard model is sufficient for the questions asked, or ideally, testing whether the sign distinction survives under alternative pRF model specifications.

(4) Interpreting resting-state transients as top-down vs. bottom-up

The event-triggered analysis labels high-amplitude DN pRF activations as "top-down events" and dATN activations as "bottom-up events." This is a reasonable inference given experience-sampling studies showing that rest involves alternation between internal and external attention, but it remains an inference. Without concurrent experience sampling, eye-tracking, or physiological monitoring, we cannot establish that a spontaneous DN transient reflects memory retrieval or internally-directed thought rather than a global arousal fluctuation. Similarly, dATN transients during rest could reflect covert shifts of spatial attention to remembered or imagined locations rather than bottom-up processing per se. I would ask the authors to soften this framing or to discuss what additional data would be needed to validate the top-down/bottom-up attribution.

(5) The "retinotopic code" vs. "visual field bias" distinction

The paper uses the language of a "retinotopic code" throughout and correctly distinguishes this from a "retinotopic map," noting that DN voxels do not form a continuous topographic representation on the cortical surface. This distinction deserves greater emphasis. In vision science, retinotopic maps carry computational significance through their topographic continuity and relationship to cortical wiring. A distributed collection of voxels with coarse visual field preferences but no cortical topography is a fundamentally different organizational feature. Recent reviews have drawn an explicit distinction between *retinotopic maps* and *visual field biases* (Groen, Dekker, Knapen & Silson, TiCS 2022), and the present findings may be more accurately characterized as the latter. Perhaps the authors think that the distinction is merely a signal-to-noise distinction, in which case I would invite them to clearly speak to this interpretation. In any case, this is not a criticism of the findings themselves, but clarity on this point would prevent conflation of two different organizational principles and would help position the work for both the vision and network neuroscience communities.

Reviewer #2 (Public review):

Summary:

Using a public dataset of retinotopic mapping and resting-state data, the authors find that the default mode network has voxels that respond (positively or negatively) to visual stimulation at specific retinotopic positions, and that resting-state activity in these voxels is correlated with activity in more traditional sensory voxels with the same visual-location preference. The retinotopic specificity is bidirectional, such that high activity in default mode voxels drives activity only in voxels with matching receptive fields in sensory cortex, and vice versa. These findings are at odds with traditional views of the default mode network as having abstract (non-retinotopic) representations and competing (rather than cooperating) with external sensory representations.

Strengths:

This study continues an intriguing line of research about how default mode regions interact with the sensory cortex. Demonstrating that there are structured interactions between these regions at rest, and that these interactions are in fact organized according to retinotopic location (as opposed to traditional views of representational format in the default mode network), provides a new framework for thinking about large-scale internal and external brain networks. The authors make use of a well-powered public dataset that allows for precise estimates of pRFs and individual-specific resting-state networks, and develop a number of interesting analyses that characterize the relationships between DN and dATN voxels. The findings are exciting and could have a major impact on future studies in cognitive neuroimaging.

The authors mention that these findings could shed light on internal/external interactions such as "anticipatory saccades or memory-guided attention," which is true, though I would argue that constructing DN representations of external stimuli is in fact even more fundamental than these specific cases (e.g., see Barnett and Bellana, 2025, "Situation models and the default mode network"). The "highways" identified in this study could play a vital role in real-world perceptual processes that are constantly translating external input into internal mental models.

Weaknesses:

(1) The criterion used for defining voxels as retinotopic seems very liberal. The authors show that only 5% of voxels have R^2>0.14 in a null analysis, and therefore define voxels with R^2>0.14 as retinotopic. Although all the networks in 1C show voxel distributions that differ from the null, the number of false positives above R^2>0.14 seems problematic, especially for the DN positive pRFs (red distribution) and to a lesser extent the DN negative pRFs (blue distribution). From visual inspection of the plot, the false discovery rate (fraction of voxels labeled as retinotopic that are false positives) looks like it would be greater than 50% for the DN-positive pRFs. The authors do show that the positive pRF voxels have above-chance consistency across runs, again providing evidence that there are true positive voxels in this set, but perhaps a stricter criterion (such as having consistent negative fits across runs) would provide more targeted identification of the DN voxels with true retinotopic sensitivity.

(2) The claim that "opponency at rest between the DN and dATN appears to be driven by the subset of DN voxels with negative retinotopic tuning" is not well supported. The fraction of DN voxels with negative pRFs is small: 9.42% of DN voxels have pRFs, and 58.77% are negative, so about 6% of DN voxels have negative pRFs. The fact that any DN voxels have negative pRFs is notable, but the authors do not provide evidence that these 6% are driving the overall behavior of the DN. They do show (e.g., in Figure 2B) that negative and positive pRFs have opposing influences, but the overall correlation with dATN does not look similar to the negative pRF connectivity. I'm also unsure whether "opponency" is a reasonable description for two networks that are "independent (i.e., not correlated)" in this analysis.

(3) The event-triggered analysis is effective at testing the bidirectional relationship between DN and dATN, with high activity in either network triggering a response in the other network. However, it would be helpful to show more validation that these "events" are meaningful windows of time to study. First, is 13 TRs a typical length of time that activity is elevated during one of these events? Second, the top-down and bottom-up terminology is perhaps too loaded and not well-justified; if the negative pRFs in the DN reflect a meaningful coding system, then couldn't low (rather than high) activity indicate a top-down event?

(4) The framing of this paper relative to the authors' past week, such as Steel et al. 2024 ("A retinotopic code structures the interaction between perception and memory systems"), could be improved. The existence of negative pRFs in the DN and a functional relationship between these pRFs and the sensory pRFs have already been described in prior work. My understanding of the primary novelty here is that this paper examines resting-state data, showing that there are widespread spontaneous interactions between broad internal and external networks, but this distinction is not made explicit in the Introduction.

(5) The definition of the default mode (DN) in this study aligns with past research, but the definition of the dorsal attention network (dATN) seems at odds with standard terminology. For example, the authors cite Fox et al. 2006, which depicts the dATN as including regions such as IPS, FEF, SMA, and MT+. Here, however, the "dATN" seems to be primarily lateral and ventral visual cortex (e.g., Figure S5). The exact location of these sensory pRFs is not critical to the authors' claims, but this labeling seems incorrect, and the motivation for defining/selecting the sensory network in this way is not described.

Reviewer #3 (Public review):

Summary:

This paper addresses an important question (the relationship between DN and dATN, and the role of retinotopic coding) and uses a set of novel analyses.

Strengths:

Important question, novel analytical approaches (pRF-informed functional connectivity analysis).

Weaknesses:

Some of the key claims are not fully supported by the data presented. There is also a concern about over-interpretation of the results. Key issues:

(1) The authors claim that retinotopic coding scaffolds the interaction between DMN and dATN. However, retinotopically tuned voxels account for a mere 9% of DMN voxels. So this appears to be a major overstatement. For instance, the statement that "these findings would position retinotopy as a unifying framework for brain-wide information processing" is not justified given the presented data.

(2) Given that positive pRF voxels in DMN positively correlate with dATN voxels and negative pRF voxels in DMN negatively correlate with dATN voxels, there is a concern that these results could be contributed to by imprecise brain network parcellations. E.g., could some of the positive pRF voxels in DMN be erroneously assigned to DMN and actually belong to one of the other task-positive networks? There is insufficient validation of network parcellation to put this worry to rest, especially since it depends on ICA, which has a degree of arbitrariness built in.

(3) The claim that retinotopic coding is intrinsic to the DN network is not supported by rigorous analysis and results. The analysis here has many arbitrary factors, including: the threshold of the 99th percentile of resting-state distribution; the designation of DN as "top-down" and dATN as "bottom-up"; the definition of "anti-matched" voxels instead of using randomly selected voxels; and the statistics being paired between matched and anti-matched voxels instead of using comparisons to baseline. Overall, I do not think that the result supports the conclusion that retinotopic coding in DN is intrinsic instead of being bottom-up-driven, given the very high threshold (99%) used and the fact that many other networks could also send bottom-up input to DN. Furthermore, the idea that bottom-up inputs only occur when the dATN (or any other RSN)'s spontaneous BOLD activity is above a certain threshold is a huge and unvalidated assumption.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

This paper leverages 7T fMRI data from the Natural Scenes Dataset to investigate whether retinotopic coding, the position-selective organization of visual response structures, spontaneous resting-state interactions between the Default Network (DN) and the Dorsal Attention Network (dATN). Using individualized network parcellations and population receptive field (pRF) modeling, the authors show that DN voxels can be split into two subpopulations based on their response to visual stimulation: those with position-specific positive BOLD responses (+pRFs) and those with position-specific negative BOLD responses (-pRFs). Critically, these subpopulations relate differently to the dATN during rest: -pRFs are anticorrelated with the dATN, +pRFs are positively correlated, and non-retinotopic DN voxels show no coupling. The anticorrelation (and positive correlation) is enhanced when DN and dATN voxels share visual field preferences. An eventtriggered analysis suggests that retinotopic coding shapes both "top-down" (DNinitiated) and "bottom-up" (dATN-initiated) spontaneous activity transients, supporting the claim that the retinotopic scaffold is intrinsic to the DN. These findings challenge the prevailing view of global DN-dATN antagonism and suggest retinotopic coding as an organizing principle for cross-network communication.

Strengths:

The central finding that what looks like network-level independence between DN and dATN decomposes into structured, bivalent interactions organized by voxellevel visual field preferences is a compelling demonstration that macro-scale network descriptions can hide meaningful substructure. The logic of the analysis is clean: pRF properties are estimated from retinotopic mapping data and then used to predict resting-state coupling in completely independent scanning sessions. This cross-session, cross-modality design rules out many circularity concerns.

The use of individualized multi-session hierarchical Bayesian parcellation (Kong et al.) to define DN and dATN boundaries within each subject is the right methodological choice for this question. Network boundaries in posterior cortex, where DN and dATN interdigitate most closely, vary considerably across individuals, and group-average approaches would introduce exactly the kind of misassignment that would most confound the result.

The matched-vs-random pRF analysis is well-controlled. The authors demonstrate that cortical distance between matched and randomly-matched dATN pRFs does not differ, effectively ruling out spatial proximity on the cortical surface as a confound. tSNR controls further show that signal quality differences do not drive the effect.

The event-triggered analysis (Figure 3) is creative and adds genuine value. Showing that retinotopically-specific coupling persists during DN-initiated activity transients, not only dATN-initiated ones, is the key piece of evidence for the claim that the code is intrinsic to the DN rather than passively inherited through bottom-up visual drive.

The result is observed consistently across all individual participants, which provides strong evidence for the robustness of the qualitative pattern despite the small sample size inherent to densely-sampled designs.

Weaknesses

(1) The nature of negative pRFs requires more scrutiny

The entire interpretive framework depends on treating negative pRFs in the DN as genuine position-selective neural responses (suppression). However, negative BOLD signals are well known to arise from non-neural sources, specifically, vascular stealing (where activation in nearby tissue diverts blood from adjacent voxels) and macrovascular draining vein effects that produce spatially displaced signal inversions. These concerns are amplified at 7T, where T2*-weighted GEEPI carries substantial macrovascular weighting. The DN and dATN interdigitate extensively in the posterior cortex, often within millimeters. A negative pRF in a DN voxel adjacent to a positive dATN voxel could, in principle, reflect the hemodynamic shadow of its neighbor rather than an independent neural response.

The spatial dispersion control (matched vs. random pRFs have similar cortical distribution) is valuable but addresses long-range confounds, not local hemodynamic crosstalk. The reliability of sign and center position across runs is reassuring but does not exclude a vascular origin, as vascular architecture is itself stable across sessions. I would encourage the authors to test whether the matched-vs-random effect survives exclusion of voxels near large pial vessels (identifiable from T2* contrast or the venograms available in the NSD). These analyses would not be dispositive, but they would meaningfully strengthen the neural interpretation.

The reviewer raises an important concern about the interpretation of negative pRFs in the DN, namely that spatially specific negative BOLD responses could, in principle, reflect local vascular effects rather than genuine position-selective suppression. The reviewer suggests excluding voxels near large vessels to address this issue.

Based on the reviewer’s suggestion, we repeated the pRF matching analysis excluding any voxels within 3mm of a major vein, as identified using the time-of-flight (TOF) MR venography included in the NSD. This analysis therefore tests whether the retinotopically specific DN–dATN coupling persists after removing voxels most likely to be affected by vascular signal.

Excluding these voxels did not impact our results: we found preferential coupling according to response valence and center position, with stronger correlation between matched +DN and +dATN voxels (t(6) = 6.054, p < 0.001), and a more pronounced negative correlation between matched -DN and -dATN voxels (t(6) = -5.0448, p < 0.01). We have added these results to the supplemental figures (Fig. S7), and also added to the text (Pg. 8). Together with the run-wise reliability of pRF sign and position, and the persistence of the matched-versus-random effect after vessel exclusion, this analysis supports the interpretation that negative DN pRFs reflect structured, spatially specific responses rather than a vascular artifact.

“Finally, to rule out any possible influences from vascular stealing (i.e. the shunting of blood into active tissue from nearby regions), we repeated the matching analysis after excluding any voxels within a 3mm radius of a major vessel (Fig. S7; see Methods). Both matching effects remained after excluding vascularly susceptible voxels (+DN x dATN: t(6) = 6.054; p < 0.001; -DN x dATN: t(6) = -5.0448; p < 0.01).”

(2) Amount of retinotopic mapping data and choice of pRF pipeline

The NSD includes 6 runs of retinotopic mapping (~5 minutes each; 3 baraperture, 3 wedge/ring). The authors use only the 3 bar-aperture runs (~15 minutes total per subject) and fit their own pRFs using AFNI's 3dNLfim procedure, rather than using the pRF estimates provided as part of the NSD release (which were fitted using the analyzePRF toolbox with all 6 runs).

Fifteen minutes of bar data is quite limited for reliable voxel-wise pRF estimation, especially in regions far from the early visual cortex, where signal-to-noise is inherently lower. Standard recommendations for robust pRF mapping in higherorder regions generally suggest substantially more data. The variance-explained threshold is close to the noise floor by design, meaning that a non-trivial number of the "retinotopic" DN voxels may be poorly estimated. Given that the core analyses depend on both the sign and the center position of these pRFs, the limited data is a significant concern.

The authors do not explain why they chose to re-fit pRFs rather than use the NSD-provided estimates. If the motivation was methodological (e.g., the NSD pRF pipeline does not readily yield signed amplitude, or the bar-only fits were judged more appropriate for detecting negative responses), this should be made explicit. If the NSD-provided pRFs can reproduce the key findings, this would substantially increase confidence in the results. If they cannot, that divergence itself would be important to understand. I would ask the authors to address this choice and, if feasible, to report whether the core results replicate using the NSDprovided pRF estimates and/or whether using all 6 runs of retinotopy data changes the findings.

The reviewer raises two related concerns: first, that the amount of retinotopic mapping data available in the NSD may be limited for estimating voxel-wise pRFs in higher-order cortical regions; and second, that we re-fit the pRF model using AFNI rather than relying on the pRF estimates provided with the NSD release. We appreciate the opportunity to clarify both points. We agree with the reviewer that more travelling bar data would be preferable and would likely yield more robust model fits, particularly in higher-order regions with lower SNR. This is a limitation of our paper that we now acknowledge in the discussion section. However, we do not think that more data would fundamentally change the pattern of our results for the following reasons.

First, we implemented a novel data-driven approach to derive a threshold for thresholding significant pRF fits (a noise floor). Importantly, our noise floor estimation yields a conservative threshold (R2 > 0.14), which is greater than both our previous work characterizing cortical pRFs (Steel et al. 2024: R2 > 0.08) and other work exploring visual responses in the default network (Klink et al. 2021: R2 > 0.05; no threshold: Szinte and Knapen 2020; Knapen 2021).

Second, the key pRF features used in our analyses – response sign and centre position – were reliable across retinotopic mapping runs. This reliability is important because our central matching analysis depends on voxel-wise estimates of both response valence and visual-field position.

Third, the matching analysis asks whether pRF parameters estimated from the retinotopic mapping task predict functional coupling measured during independent resting-state scans. Noisy or unstable pRF estimates should weaken this relationship, because they would degrade the accuracy of voxel-wise matching. Thus, parameter instability would be expected to obscure retinotopically specific coupling rather than systematically produce the observed matched-versus-random effects.

To the reviewer’s question about our decision to re-fit the pRF model using AFNI, the reviewer is correct that this was motivated by the requirements of our analysis: we re-fit the pRF estimates using AFNI because it allows for both positive and negative signed amplitudes. The pRF model fits provided with the NSD do not allow bivalent amplitude estimates. We have made this decision clearer in the text, reproduced below (Pg. 4-5; Pg. 14-15).

“We chose to re-fit the data using a simple Gaussian approach as implemented in AFNI to allow for both positive and negative signed amplitudes.”

“The limited amount of pRF mapping task data included in the NSD posed a challenge for establishing reliable visual response estimates. Here, we addressed this issue by developing a novel thresholding method to establish robust voxel-wise model fits. Among voxels that passed this empirical threshold, we observed a significant correlation in voxel-wise estimates of centre position and visual response amplitude. In addition, our pRF matching results were based on the relationship between the voxel-wise estimates of centre position and response amplitude with resting-state fMRI – a completely independent measure. Crucially, noisy estimates of pRF parameters would obscure this relationship and make our results less likely. Therefore, despite the relatively limited pRF mapping data available, unstable pRF estimates are unlikely to drive our results.”

(3) pRF model adequacy for the Default Network

The isotropic Gaussian pRF model was developed for and validated in early and mid-level visual cortex, where it captures the dominant spatial selectivity of neuronal populations. In DN voxels where the model explains comparatively little variance, it is less clear that the model is capturing the right quantity.

Specifically, the negative pRFs could conceivably be described by a model with a dominant suppressive surround (e.g., a difference-of-Gaussians model), in which what appears as a "negative pRF" in the standard model is actually the surround component of a center-surround mechanism whose center is poorly resolved. This distinction matters: a genuine inverted code (negative center response) implies a qualitatively different computation than inherited surround suppression from nearby visual cortex.

The authors should consider discussing why the standard model is sufficient for the questions asked, or ideally, testing whether the sign distinction survives under alternative pRF model specifications.

We appreciate the reviewer’s comment about the limitations of a single gaussian pRF model. We chose the single gaussian model as a direct extension of prior work from our lab and others (Steel et al., 2024, Klink et al., 2022, Szinte and Knapen, 2021). We agree that a negative response in this model could, in principle, reflect a more complex spatial profile, such as a dominant suppressive surround. However, adjudicating among alternative pRF models would require more retinotopic mapping data than are available in the NSD, particularly for higher-order cortex. Thus, we feel that it is outside the scope of the current work. We now address this limitation in our discussion (Pg. 15).

“Relatedly, here we used a single gaussian model, consistent with prior work on negative visual responses in memory systems (31, 33, 34). However, other models of visual receptive fields might offer further insight into the DN’s visual responsiveness, such as double gaussian models of surround suppression (65) or compressive summation (66). Future studies might directly compare different visual models to further refine the computations underpinning visual responses in the DN.”

(4) Interpreting resting-state transients as top-down vs. bottom-up The event-triggered analysis labels high-amplitude DN pRF activations as "topdown events" and dATN activations as "bottom-up events." This is a reasonable inference given experience-sampling studies showing that rest involves alternation between internal and external attention, but it remains an inference. Without concurrent experience sampling, eye-tracking, or physiological monitoring, we cannot establish that a spontaneous DN transient reflects memory retrieval or internally-directed thought rather than a global arousal fluctuation. Similarly, dATN transients during rest could reflect covert shifts of spatial attention to remembered or imagined locations rather than bottom-up processing per se. I would ask the authors to soften this framing or to discuss what additional data would be needed to validate the top-down/bottom-up attribution.

The reviewer raises an important concern about the strong interpretation of elevated BOLD activity detected in the DN and dATN as top-down and bottom-up events. We agree that the limitations of fMRI in our current data prevent these strong claims about the origin of these signals. We have therefore softened this framing throughout the manuscript, and we now refer to these events as DN-driven and dATN-driven. We think that this more directly describes the analysis: events were defined by transient high-amplitude activity in DN or dATN pRFs, respectively.

(5) The "retinotopic code" vs. "visual field bias" distinction The paper uses the language of a "retinotopic code" throughout and correctly distinguishes this from a "retinotopic map," noting that DN voxels do not form a continuous topographic representation on the cortical surface. This distinction deserves greater emphasis. In vision science, retinotopic maps carry computational significance through their topographic continuity and relationship to cortical wiring. A distributed collection of voxels with coarse visual field preferences but no cortical topography is a fundamentally different organizational feature. Recent reviews have drawn an explicit distinction between retinotopic maps and visual field biases (Groen, Dekker, Knapen & Silson, TiCS 2022), and the present findings may be more accurately characterized as the latter. Perhaps the authors think that the distinction is merely a signal-to-noise distinction, in which case I would invite them to clearly speak to this interpretation. In any case, this is not a criticism of the findings themselves, but clarity on this point would prevent conflation of two different organizational principles and would help position the work for both the vision and network neuroscience communities.

The reviewer raises a valuable point about the distinction between a retinotopic code, a retinotopic map, and a visual field bias, and we are happy to add discussion of this topic to our manuscript.

Our results show that the DN does not exhibit a continuous retinotopic map in the sense used in early visual cortex. Rather, our results suggest a distributed voxel-level code for visual-field position: individual DN voxels show reliable spatial preferences, and these preferences predict retinotopically specific functional coupling with dATN voxels. This voxel-level organization is analogous to other distributed spatial codes, such as head-direction coding in retrosplenial cortex, where spatial variables are represented by population activity without requiring a topographic map on the cortical surface. This differs from a coarse visual-field bias, including preferential responses to the contralateral visual field, although we do also observe such biases. We have added text unpacking this important distinction to the Discussion (Pg. 15-16):

“Prior work has emphasized the visual response bias in regions where voxel-wise retinotopic responses lack a map-like organization(35); overall, the DN does exhibit this kind of bias. However, our results show that the voxel-scale activity underpinning this bias reflects the latent connectivity of those voxels. Thus, we adopt the term “retinotopic coding”, because this voxel-scale coding scheme exists without a map-like organization on the cortical surface. For example, rodent and bat head direction cells are not laid out in a literal ring, but the population code of these neurons forms a ring manifold(68, 69).”

Reviewer #2 (Public review):

Summary:

Using a public dataset of retinotopic mapping and resting-state data, the authors find that the default mode network has voxels that respond (positively or negatively) to visual stimulation at specific retinotopic positions, and that restingstate activity in these voxels is correlated with activity in more traditional sensory voxels with the same visual-location preference. The retinotopic specificity is bidirectional, such that high activity in default mode voxels drives activity only in voxels with matching receptive fields in sensory cortex, and vice versa. These findings are at odds with traditional views of the default mode network as having abstract (non-retinotopic) representations and competing (rather than cooperating) with external sensory representations.

Strengths:

This study continues an intriguing line of research about how default mode regions interact with the sensory cortex. Demonstrating that there are structured interactions between these regions at rest, and that these interactions are in fact organized according to retinotopic location (as opposed to traditional views of representational format in the default mode network), provides a new framework for thinking about large-scale internal and external brain networks. The authors make use of a well-powered public dataset that allows for precise estimates of pRFs and individual-specific resting-state networks, and develop a number of interesting analyses that characterize the relationships between DN and dATN voxels. The findings are exciting and could have a major impact on future studies in cognitive neuroimaging.

The authors mention that these findings could shed light on internal/external interactions such as "anticipatory saccades or memory-guided attention," which is true, though I would argue that constructing DN representations of external stimuli is in fact even more fundamental than these specific cases (e.g., see Barnett and Bellana, 2025, "Situation models and the default mode network"). The "highways" identified in this study could play a vital role in real-world perceptual processes that are constantly translating external input into internal mental models.

Weaknesses:

(1) The criterion used for defining voxels as retinotopic seems very liberal. The authors show that only 5% of voxels have R^2>0.14 in a null analysis, and therefore define voxels with R^2>0.14 as retinotopic. Although all the networks in 1C show voxel distributions that differ from the null, the number of false positives above R^2>0.14 seems problematic, especially for the DN positive pRFs (red distribution) and to a lesser extent the DN negative pRFs (blue distribution). From visual inspection of the plot, the false discovery rate (fraction of voxels labeled as retinotopic that are false positives) looks like it would be greater than 50% for the DN-positive pRFs. The authors do show that the positive pRF voxels have abovechance consistency across runs, again providing evidence that there are true positive voxels in this set, but perhaps a stricter criterion (such as having consistent negative fits across runs) would provide more targeted identification of the DN voxels with true retinotopic sensitivity.

We thank the reviewer for giving us the opportunity to discuss this important decision. We agree with the reviewer that a stricter R2 criterion could result in more targeted pRF identification. Motivated by the reviewer’s suggestion, we repeated the cross-region pRF matching analysis across multiple R2 thresholds.

The retinotopic matching effects were not dependent on the original threshold. In fact, we found that the pRF matching effects are enhanced as the R2 value increases (Fig. S5). This pattern suggests that any false-positive voxels admitted near the original threshold would dilute, rather than drive, the observed matched-versus-random effects. We have added text to the results highlighting this finding (Pg. 7):

“In contrast, DN voxels that responded positively to visual stimulation (DN positive pRFs, +pRFs) had a positive correlation with the dATN (mean correlation = 0.284±0.152, t(6) = 4.96, p = 0.0025), while DN voxels with systematic negative responses to visual stimulation (DN negative pRFs, -pRFs) were anti-correlated with the dATN (mean correlation = -0.21±0.149, t(6) = -3.75, p = 0.0094). This relationship was further strengthened by adopting more conservative R2 thresholds up to 0.30 despite the overall number of included voxels decreasing, suggesting that this effect is not driven by false-positive voxels at the edge of our threshold criteria (Fig. S5).”

(2) The claim that "opponency at rest between the DN and dATN appears to be driven by the subset of DN voxels with negative retinotopic tuning" is not well supported. The fraction of DN voxels with negative pRFs is small: 9.42% of DN voxels have pRFs, and 58.77% are negative, so about 6% of DN voxels have negative pRFs. The fact that any DN voxels have negative pRFs is notable, but the authors do not provide evidence that these 6% are driving the overall behavior of the DN. They do show (e.g., in Figure 2B) that negative and positive pRFs have opposing influences, but the overall correlation with dATN does not look similar to the negative pRF connectivity. I'm also unsure whether "opponency" is a reasonable description for two networks that are "independent (i.e., not correlated)" in this analysis.

The reviewer raises an important point about whether negative DN pRFs should be described as driving the overall DN–dATN relationship. We agree that this language was too strong. Negative pRFs constitute a small subset of DN voxels, and our analyses show that this subset has a distinct pattern of functional coupling with the dATN, not that it explains the global relationship between the DN and dATN as a whole.

We have therefore revised the manuscript to avoid implying that negative DN pRFs drive overall DN–dATN opponency. Instead, we now frame these voxels as an important retinotopically tuned subpopulation nested within broader network dynamics. Specifically, our results show that visually responsive DN voxels are not homogeneous: positive and negative DN pRFs show opposing patterns of coupling with dATN pRFs, and these interactions are strengthened when voxels share visual-field preferences. This suggests that a small but structured subset of DN voxels may provide a route for retinotopically specific communication between internally and externally oriented networks, without implying that this subset determines the mean activity pattern of the entire DN:

“Spontaneous DN and dATN activity during rest is uncorrelated at the network level. However, voxel-scale functional coupling across networks is shaped by the latent visual field preferences of individual voxels in each network, as measured during independent retinotopic mapping.” Abstract (Pg. 2)

“This result shows that voxel-level visual response profiles shape DN-dATN coupling during spontaneous resting-state activity. Specifically, the DN and dATN activation is independent during rest. However, at the voxel-level, specific sub-groups of DN voxels have distinct coupling patterns with the dATN that depends on the valence of voxels’ visual responses. DN and dATN voxels with positive visual responses show a positive relationship during rest, and a notable subset of DN voxels with negative visual responses display the canonical opponency with dATN voxels. This suggests that retinotopic coding may be a mechanism that enables visual information to be exchanged between these large-scale brain systems. Specifically, opponency at rest between the DN and dATN appears to be driven by the subset of DN voxels with negative retinotopic tuning.” Results (Pg. 7)

These findings offer a multi-scale account of neural communication, in which interactions among sub-populations of voxels with shared tuning preferences are nested within macro-scale network dynamics. Nesting multiple neural codes might enable ongoing computations within a larger brain system (e.g., attending to internal mental states within the DN during memory recall), while simultaneously allowing for the sharing of fine-grained representations across brain systems (34). Discussion (Pg. 14)

(3) The event-triggered analysis is effective at testing the bidirectional relationship between DN and dATN, with high activity in either network triggering a response in the other network. However, it would be helpful to show more validation that these "events" are meaningful windows of time to study. First, is 13 TRs a typical length of time that activity is elevated during one of these events? Second, the top-down and bottom-up terminology is perhaps too loaded and not well-justified; if the negative pRFs in the DN reflect a meaningful coding system, then couldn't low (rather than high) activity indicate a top-down event?

We thank the reviewer for these helpful suggestions. To the best of our knowledge, there is not currently a widely agreed-upon time window for performing event-based fMRI analyses. We chose a 13 TR time window to balance between sufficiently capturing BOLD signal related to the chosen event while also minimizing influence from other signal fluctuations, based on the procedure adopted in Gordon et al. (Nature, 2023) and Mitra et al. (J. Neuro Phys, 2014), which considered temporal relationships among brain regions over comparable timescales. In our analysis, this window considered 6 TRs (9.6s) on either side of the detected event, which we felt comfortably captures the peak BOLD signal that would result from an impulse at the event time, and responses that may reflect upstream activity leading into it.

The reviewer has raised an additional comment about the terms “top-down” and “bottom-up.” These concerns were shared by Reviewer 1. Based on these comments, we have adopted the terms “DN-driven” and “dATN-driven”, which we think aligns more closely with our analysis approach.

(4) The framing of this paper relative to the authors' past work, such as Steel et al. 2024 ("A retinotopic code structures the interaction between perception and memory systems"), could be improved. The existence of negative pRFs in the DN and a functional relationship between these pRFs and the sensory pRFs have already been described in prior work. My understanding of the primary novelty here is that this paper examines resting-state data, showing that there are widespread spontaneous interactions between broad internal and external networks, but this distinction is not made explicit in the Introduction.

We appreciate the opportunity to clarify the novel aspects of our paper. The reviewer correctly identifies the extent of prior work, which identified -pRFs in regions of the canonical default network (Szinte and Knapen, 2021; Klink et al., 2022) and characterized the local interactions between adjacent perceptual and mnemonic regions (Steel et al., 2024). Our current work builds upon these findings in two key ways.

First, we explore the effect across individually-defined whole brain networks. While the DN and dATN are often adjacent, these networks are spatially discontinuous and are comprised of distinct sub-regions (e.g. in prefrontal cortex). Whether retinotopic patterning of activity would persist in distributed networks could not have been extrapolated from our prior work. We think that finding will be of broad interest to the community studying perception and memory systems, because it offers a mechanistic account of how information is read in/out of memory.

Second, here we considered whether spontaneous activity across networks would be structured by a retinotopic code. Our previous work characterized activity during tasks that depended on visual information: either scene perception or mental imagery. While the prior work was an important first step, it left open the possibility that retinotopic coding may only be relevant in visual tasks. By demonstrating that the retinotopic coding structures voxel-specific coactivation during rest, which entails no overt visual demands, we provide evidence that retinotopic features are a general, mode-agnostic code between regions.

(5) The definition of the default mode (DN) in this study aligns with past research, but the definition of the dorsal attention network (dATN) seems at odds with standard terminology. For example, the authors cite Fox et al. 2006, which depicts the dATN as including regions such as IPS, FEF, SMA, and MT+. Here, however, the "dATN" seems to be primarily lateral and ventral visual cortex (e.g., Figure S5). The exact location of these sensory pRFs is not critical to the authors' claims, but this labeling seems incorrect, and the motivation for defining/selecting the sensory network in this way is not described.

We thank the reviewer for this insightful comment and their careful consideration of our network definition.

Our method for network identification, and the topography of the resulting networks, are broadly consistent with more recent conceptualizations of the DN and dATN (e.g. Du et al. 2024, Gordon et al. 2017, Braga and Buckner 2017). Relatedly, because we defined brain networks based on the unique connectivity patterns of each individual participant, we expect them to differ from previous group-level network descriptions. The increased resolution of the 7T data in the NSD may also result in greater departure from prior definitions compared to previous work done at 3T.

Further study into dATN differences between group-level 3T, individualized 3T, and individualized 7T networks could be a valuable future direction, but this is outside the scope of this work.

Reviewer #3 (Public review):

Summary:

This paper addresses an important question (the relationship between DN and dATN, and the role of retinotopic coding) and uses a set of novel analyses.

Strengths:

Important question, novel analytical approaches (pRF-informed functional connectivity analysis).

Weaknesses:

Some of the key claims are not fully supported by the data presented. There is also a concern about over-interpretation of the results. Key issues:

(1) The authors claim that retinotopic coding scaffolds the interaction between DMN and dATN. However, retinotopically tuned voxels account for a mere 9% of DMN voxels. So this appears to be a major overstatement. For instance, the statement that "these findings would position retinotopy as a unifying framework for brain-wide information processing" is not justified given the presented data.

We appreciate the reviewer’s concern about the framing of our conclusion, which was shared by reviewers 1 and 2. In response to these comments, we have revised our paper to more accurately reflect the observed data. Specifically, we focus on the specific sub-populations of voxels within the DN and dATN that show retinotopic responses, and we have removed references to explaining the overall pattern of activity across networks.

(2) Given that positive pRF voxels in DMN positively correlate with dATN voxels and negative pRF voxels in DMN negatively correlate with dATN voxels, there is a concern that these results could be contributed to by imprecise brain network parcellations. E.g., could some of the positive pRF voxels in DMN be erroneously assigned to DMN and actually belong to one of the other task-positive networks? There is insufficient validation of network parcellation to put this worry to rest, especially since it depends on ICA, which has a degree of arbitrariness built in.

We thank the reviewer for the opportunity to clarify our method for network definition.

Precision functional mapping is a growing field with many methods for defining personalized functional networks for each individual. Because the NSD resting-state data is relatively high resolution, we chose an approach designed to improve the stability of voxel-wise network assignment: Multi-Session Hierarchical Bayesian Modeling approach (Kong et al. 2019; Du et al. 2024). This approach enhances stability of network assignment by including a group-based prior and accounting for both within- and across-subject variability. This approach is more stable than ICA, and, because this approach leverages a prior, there is less concern about arbitrary or idiosyncratic network definitions.

However, it is still common for network assignments to have lower confidence around the borders between networks. Yet, we also do not think border misassignment is likely to explain the present results for two reasons: first, while DN and dATN nodes are sometimes adjacent, there are many regions where they are spatially distant, such as the IPS for dATN and the lateral temporal lobe for DN. Second, the DN pRFs do not appear to cluster selectively along DN–dATN borders, suggesting that they are not simply misassigned dATN voxels.(Fig. S3) Therefore, we think voxels on the edge of these networks are unlikely to drive the effects observed here (see Fig. S3).

(3) The claim that retinotopic coding is intrinsic to the DN network is not supported by rigorous analysis and results. The analysis here has many arbitrary factors, including: the threshold of the 99th percentile of resting-state distribution; the designation of DN as "top-down" and dATN as "bottom-up"; the definition of "anti-matched" voxels instead of using randomly selected voxels; and the statistics being paired between matched and anti-matched voxels instead of using comparisons to baseline. Overall, I do not think that the result supports the conclusion that retinotopic coding in DN is intrinsic instead of being bottomup-driven, given the very high threshold (99%) used and the fact that many other networks could also send bottom-up input to DN. Furthermore, the idea that bottom-up inputs only occur when the dATN (or any other RSN)'s spontaneous BOLD activity is above a certain threshold is a huge and unvalidated assumption.

The reviewer raises several interesting concerns about decisions in our event-detection analysis. Here, we clarify the rationale for several analytic choices:

(1) The 99th-percentile threshold was chosen to identify sparse, high-amplitude events while minimizing contamination from smaller ongoing fluctuations.

(2) The other reviewers also noted a concern with the top-down/bottom-up terminology. We have revised these terms to DN-driven and dATN-driven, which we think reflect our approach more accurately.

(3) We used anti-matched rather than randomly selected voxels because the full event-by-voxel randomization procedure was computationally intractable at the network level.

(4) We did not understand the reviewer’s contention about activation baseline, but we would welcome clarification.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Minor points

(1) The reliability analysis (Figure 1D) notes that dATN negative pRF amplitude was not reliable above chance in 2 of 7 participants. This could be discussed more prominently as it suggests that negative pRFs may not be stable features in all networks, which tempers the generality of the sign distinction as a fundamental organizational property.

We thank the reviewer for raising this point of clarification. It is true that 2/7 participants did not show reliable negative pRFs in the dATN. However, the majority of participants show stable negative pRFs, and even in these 2 participants, the negative result does not indicate that negative pRFs would not be stable in those individuals with additional data.

Based on the reviewer’s comment, we have added emphasis to this point, but we do not feel that this warrants greater discussion in the paper.

Both positive and negative pRF amplitude was reliable in the DN for all subjects. In the dATN, positive amplitude pRFs were reliable in all participants, and negative amplitude pRFs (which constituted a small proportion of the overall pRFs in this network) were reliable in 5/7 participants. For the remainder of the paper, we only consider positive pRFs in the dATN. Importantly, pRF center position was highly reproducible across runs of pRF data in the dATN and DN in all subjects (Fig. 1D). Pg. 5

(2) The paper would benefit from situating the findings more explicitly within the cortical gradient framework (Margulies et al., 2016), which predicts that DN regions have maximally abstract, transmodal codes. The present findings complicate this view productively and deserve to be "situated" within that ongoing debate.

We agree that the gradient framework is interesting, and we have added discussion of Margulies to our paper. (Pg. 16)

Relatedly, the DN is considered a transmodal hub for cortical processing, where disparate sensory and motor processes converge (59, 75) The DN’s position at the cortical apex implies connections with and influence over unimodal cortical areas. However, the mechanism for liking unimodal and transmodal networks had been unknown. Prior work posited that sensory coding in transmodal areas might serve this function (31, 35), and our data provide direct empirical support for this account: specific visually-responsive voxels provide an input/output interface linking perceptual and memory systems. This complements work delineating specific affective and effective subregions within the DN that link the DN to other brain areas (76). Thus, while the DN may be “distant from input” (28), these data suggest that it is not disengaged from sensory processing.

(3) It would be informative to know whether the *proportion* of negative vs. positive pRFs differs between DN-A and DN-B, given their distinct functional roles.

Despite the functional specialization of DN-A and DN-B, and the slightly higher mean proportion of negative pRFs in DN-A (61% vs 56%), we found no statistically significant difference in the proportion of negative pRFs across the two networks (t(6) = 0.888, p = 0.409).

(4) Low N is inherent to the densely-sampled NSD design, and the within-subject consistency is a strength. Nevertheless, with 6 degrees of freedom, the precision of specific quantitative estimates (e.g., that 58.77% of DN pRFs are negative) is uncertain, and the authors should be cautious about the generalizability of these point estimates.

The reviewer raises a concern about the inclusion of specific levels of decimal place in our statistical reporting. We do not think that this is a major issue with the paper, but we are willing to change if the reviewer feels strongly.

Reviewer #2 (Recommendations for the authors):

(1) Figure 1C could use an explicit legend (I believe it is following the color convention from the bar plots in 1F?). Also, for consistency, it would be helpful to make all the colormaps in 1F correspond to the bars (i.e., change the dATN colormap to go white->green).

We thank the reviewer for this suggestion, and we have added explicit labels to Fig. 1C

(2) Providing a scatter plot, in which each dot is a voxel and the x and y axes are the pRF amplitude estimates in different runs, could help provide evidence that there are voxels with pRFs that have consistently negative amplitudes across runs. This would also go beyond the binary consistency analysis in Figure 1D to show that the magnitudes of the amplitude estimates are also consistent.

We thank the reviewer for this suggestion. We feel that the binary consistency conveys sufficient information. Because the analysis is done using pairwise correlation, how the scatter plot would reflect the three-way consistency is not clear.

(3) For understanding how the overall correlation between DN and dATN could be driven by voxel populations with opposing effects (e.g., Figure 2B), it would be useful to show how the +pRF and -pRF voxels compare to other voxels within the DN. For example, are these the voxels with the strongest negative and positive correlations with dATN, or are there many other DN voxels (among the 90% that do not have pRFs) that also have similarly-strong dATN correlations?

The reviewer offers a very interesting suggestion. Based on the reviewer’s suggestion, we have refocused our paper on the particular subpopulations of +/- pRFs in the DN, rather than on an explanation for the overall pattern of correlation between the DN and dATN. Because our revised framing focuses on the properties of these retinotopically defined voxel populations, rather than on explaining whole-network DN–dATN coupling, we have not added this additional analysis. We have revised the relevant text to avoid implying that these pRF subpopulations drive the overall network-level relationship.

(4) Initially, the baseline comparison pRFs for the matched pRFs are labeled "random" pRFs, which seems misleading; these are closer to "mismatched"/"anti-matched" pRFs since they are selected from the 1/3 that are farthest away. Then the comparison switched to using the anti-matched pRFs that are the 10 very farthest away, though I didn't understand the rationale that "the large number of pRFs made the random matching procedure impractical" - in what way is the number of pRFs larger in this analysis? Having a more consistent baseline (e.g., just using the 10 anti-matched pRFs the whole time) would be easier to interpret.

We thank the reviewer for this suggestion. We have compared the results between the randomly-sampled bottom ⅓ matched versus the 10 worst matched, and the pattern of results is identical (the effect is strongest in the 10 worst matched). Therefore, we include the bottom ⅓ matched in the main text as a more conservative test of this effect. We are happy to include this as a supplemental figure if the reviewer feels it is essential.

(5) In the past, I have only seen the terminology "bootstrapped" to refer to sampling with replacement from the data sample, producing samples/statistics that are centered on the observed data. Here (lines 704-708), the sampling is coming from the null distribution of randomly-chosen voxels, and therefore the term "bootstrapped" would not apply (and could just be replaced with "null").

We have revised this terminology in the manuscript.

Reviewer #3 (Recommendations for the authors):

(1) Abstract and Discussion should be significantly toned down. E.g., the claim that "These findings challenge the prevailing view of global DN-dATN antagonism" is not really supported by the data provided. The claim that "retinotopic coding underpins the dynamic coordination of perception and thought" is also unsupported by the presented data.

We have revised the manuscript in light of this comment.

(2) Line 233-235: The null statistical result cannot support the claim reached here. Correlation analysis or Bayesian statistics should be used.

We have revised the manuscript in light of this comment.

(3) Line 250-254: Comparison to baseline should be used, in addition to comparing matched and random voxels.

We agree that baseline comparisons can be useful in event-triggered analyses. However, for the pRF-matching analysis discussed here, the critical question is whether shared visual-field preference influences resting-state functional coupling between DN and dATN voxels. For this question, we believe that the appropriate baseline is the coupling observed for pRFs that do not share visual-field preferences. We therefore compare retinotopically matched pRFs to randomly matched pRFs drawn from the same networks.

(4) Line 271: "not" is missing.

We have revised the manuscript in light of this comment.

Version 2

Abstract

The human brain seamlessly integrates internally generated thoughts with incoming sensory information, yet the large-scale networks that support these functions -- the (Default Network, DN) and external (Dorsal Attention Network, dATN) -- are traditionally viewed as functionally independent. This raises a crucial question: how does the brain integrate information across these seemingly non-interactive systems? Here, using densely sampled 7T fMRI, individualized resting-state parcellations, and voxel-wise population-receptive-field mapping, we show that these internal/external networks are more interlocked than previously thought. Spontaneous DN and dATN activity during rest is uncorrelated at the network level. However, voxel-scale functional coupling across networks is shaped by the latent visual field preferences of individual voxels in each network, as measured during independent retinotopic mapping. Voxels that share visual field preferences exhibit stronger spontaneous coupling than those with divergent preferences. These retinotopically-specific interactions are bivalent: DN voxels with negative (suppressive) visual response amplitudes are anticorrelated with matched (positive) dATN voxels, while those DN voxels with positive response amplitudes are positively correlated. Thus, distinct subpopulations of visually-tuned DN voxels participate in spatially-specific interactions with the dATN. Further, retinotopic coding is intrinsic to the DN, persisting even during periods when the DN signal is elevated. These findings reveal a latent, voxel-level architecture of retinotopically-grounded interactions between the DN and dATN. Taken together, our results suggest that retinotopic coding underpins the dynamic coordination of perception and thought in the human brain.

Main Text

A fundamental goal of neuroscience is to understand how the coordinated activity of distributed brain networks gives rise to cognition (1–9). Central to this aim is understanding principles that govern interactions among networks with distinct computational roles. In particular, it is unclear how networks involved in externally-oriented attention (e.g., processing sensory input) interact with those that support internally-oriented attention (e.g., introspection and memory) (10–14). Understanding this interplay is crucial, because the dynamic exchange between internal and external representations underlies how we perceive, remember, and navigate our world -- yet the ’common language’ that enables communication between these systems remains unclear.

Historically, the Default Network (DN) and Dorsal Attention Network (dATN)—prototypical networks for internal and external attention, respectively—have been viewed as functionally competitive (10–12, 14, 15). Classic neuroimaging studies show that externally oriented tasks activate the dATN and deactivate the DN (10, 13, 14, 16–18), while internally oriented tasks produce the opposite pattern (6, 19–23). This dissociation is also observed at rest, where DN and dATN activity is typically independent or anti-correlated (11, 12, 24, 25), depending on the application of global signal regression(11, 26). Such findings have led to the prevailing view that DN/dATN antagonism supports selective attention to either internal or external information, minimizing interference between perception and memory (10, 12, 27, 28). Yet, this framework leaves open a fundamental question: if these networks are functionally dissociated, how does the brain integrate perceptual and mnemonic information when both are required – for example during anticipatory saccades or memory-guided attention?

Recent studies of the voxel-level coding properties within these large-scale networks offer insight into this question. Although the DN is traditionally associated with abstract or semantic coding(2, 28–30), emerging evidence suggests that it also exhibits a neural code typically associated with externally-oriented visual processing: coding of retinotopic stimulus position(31–34). Notably, retinotopic coding in the DN differs from that observed in traditional visually responsive areas, including the dATN. While visual stimulation elicits position-dependent increases in neural activity in traditional visual areas, many DN voxels often show position-dependent decreases in activity(31, 33, 34). This voxel-level pattern suggests that the DN’s global deactivation during externally-oriented tasks may not reflect a simple disengagement from external input, but rather a structured, spatially-tuned response to visual stimulation.

What purpose might the DN’s voxel-level retinotopic code serve? We and others have proposed that the visual code in the DN might play a functional role in structuring mnemonic-perceptual interactions (31, 32, 34, 35). Recent work provides critical support for this theory, revealing that scene-selective memory regions near the DN(36) exhibit voxel-level retinotopic opponency with scene-selective visual areas (34). Specifically, the inverted retinotopic code in mnemonic regions is functionally tethered to the positive code in perceptual regions: during perception, visual stimuli at specific field locations activate perceptual voxels while suppressing memory voxels tuned to the same location, a pattern that reverses during memory retrieval (34). These results indicate that visual field location can coordinate dynamic interactions between specialized perceptual and mnemonic areas at the cortical apex, raising the intriguing hypothesis that retinotopic grounding might support perceptual-mnemonic interactions more broadly across the brain. Yet, whether retinotopic coding serves as a more general organizing principle for interactions across large-scale networks remains unknown.

Here, we tested the hypothesis that global interactions across large-scale internally- and externally-oriented brain networks are structured at the voxel level by a shared retinotopic code35. This framework challenges the traditional view of internally- and externally-oriented brain networks as globally antagonistic, proposing instead that voxel-level retinotopic tuning enables functional coupling and shared information processing between them. To evaluate this hypothesis, we leveraged a high-resolution 7T fMRI dataset(37), individualized network identification(3, 4, 8), and voxel-wise modeling(38) to test three core predictions. First, voxel-wise retinotopic coding should be robustly present within individuals’ DN and dATN networks, but with divergent tuning polarities: positive retinotopic responses should dominate the dATN, while DN should contain a significant proportion of retinotopic voxels with negative response amplitudes to visual stimulation. Second, we predicted that spontaneous DN–dATN interactions during resting-state fMRI would be structured by the retinotopic properties of individual voxels. Specifically, functional coupling should be stronger between voxels that share visual field preferences across networks, revealing that retinotopic position-specific responses form a latent architecture to scaffold cross-network interactions even in the absence of overt task demands. Third, we predicted that these retinotopically grounded interactions would support both coupling driven by the DN and coupling driven by dATN, suggesting that shared spatial tuning enables dynamic, bidirectional interactions. Together, these findings would position retinotopy as a unifying framework for brain-wide information processing and for the coordination of internal and external attentional dynamics.

Retinotopic coding in internally and externally oriented networks

To investigate the role of retinotopic coding in structuring interactions between internally- and externally-oriented brain networks, we analyzed 7T fMRI data from the Natural Scenes Dataset(37), acquired from the same densely-sampled participants during both resting-state and retinotopic mapping sessions (Fig. 1A-B).

Inversion of retinotopic coding between externally- and internally- oriented brain networks.

A. Population receptive field (pRF) modeling with fMRI. We established visual field preferences for each voxel using pRF modelling(38). Voxels with positive BOLD responses to the visual stimulus are classified as positive pRFs (+pRFs), and those with negative BOLD responses are classified as negative pRFs (-pRFs). B. Individualized resting-state network parcellation. Resting-state fMRI was collected in all participants (N=7; 34-102 minutes per participant) and used to derive individualized cortical network parcellations. Parcellations were generated using the multi-session hierarchical Bayesian modelling approach with the Yeo 15 HCP atlas as a prior(8, 42). C. PRF model variance explained is greater when fit to the pRF mapping task versus resting-state fMRI data for negative and positive pRFs in the DN and dATN. Histograms show voxel-wise variance explained for all pRFs fit to rest (dark) and the pRF modeling task fMRI data (light). All distributions were significantly shifted rightward for data fit to the pRF modeling task (dATN pRFs: K-S test: D-=0.655, p<0.001; Negative DN pRFs: D-=0.232, p<0.001; Positive DN pRFs: D-=0.124, p<0.001). Dashed lines indicate the 95th percentile of R2 within the null distribution used; we chose the threshold for visually-defined voxels for subsequent analyses (R2 > 0.14) to exceed these values. D. DN and dATN pRF amplitude and center location estimates are reliable. To assess reliability of pRF amplitude estimates, we compared voxel-wise signed amplitude by computing the dice coefficients of binarized pRF amplitude maps (i.e., positive or negative pRFs) between pairs of pRF modelling runs. To estimate center position, we calculated the distance between pairs of pRF centers iteratively between pRF modeling runs. For both metrics, chance was estimated by performing the same computation on a random sampling of voxels drawn from both pRFs and non-pRFs (5000 iterations for each metric). In all subjects, for dATN positive and DN positive and negative pRFs, pRF amplitude and center position exceeded what would be expected by chance (z-scored Dice/distance values > 2.9). Although only dATN positive pRFs were included in subsequent analyses, the dATN pRF amplitude was not reliable above chance in two participants; this suggests that negative amplitude pRFs are not stable in all cortical networks. E. The internally-oriented DN contains a larger proportion of negative pRFs than the externally-oriented dATN. Proportion of negative pRFs (of total pRFs that survive thresholding in that network) in all cortical networks are shown. Inset. DN-A and DN-B contain more negative pRFs than the dATN (t(6)=7.94, p<0.001), and there is no difference in the proportion of negative pRFs among sub-networks. F. DN and dATN pRFs networks show contralateral bias. Coverage plots across participants for positive dATN pRFs and positive and negative DN pRFs. Bar plots show the difference in coverage between contra- versus ipsi-lateral visual fields for each pRF type. Bar plots show mean across participants, connected points represent individual participant data.

We first identified individualized large-scale networks for each participant using their resting-state fMRI data (8, 39) (Fig. 1B, S1). Resting-state data were preprocessed using independent components analysis (ICA) with manual noise component selection(40, 41), and no global signal regression was performed(26). Networks were identified using a multi-session hierarchical Bayesian modeling approach(8, 39, 42), which leverages repeated data runs to estimate within participant variation and derive stable parcellations for 15 cortical networks, including the Default Networks A-B and the Dorsal Attention Networks A-B (Fig. 1B). Unless otherwise noted, we combined the Default Networks A-B and the Dorsal Attention Networks A-B to constitute our internal and externally oriented networks, respectively (hereafter, DN and dATN).

Next, we determined the visual field preferences of all voxels within these individualized networks using a voxel-wise population receptive field (pRF) modeling approach(38, 43) to fit each participant’s BOLD activity during retinotopic mapping(38) (Fig 1A, S2). We fit the data using a gaussian model (implemented in AFNI) that allow for both positive and negative signed amplitudes.

We considered any voxel with >14% variance explained by our pRF model to be exhibiting a retinotopic code, and hereafter we refer to these voxels as “pRFs”. We determined this threshold by defining the noise floor of our retinotopic model in the DN and dATN by comparing task-derived model fits to a control null distribution derived from resting-state fMRI data (Fig. 1C; distribution of voxel-wise variance explained by task versus null, K-S tests: D-s>0.124, ps<0.001). Two pRF features were of primary interest: 1) center position, which estimates the preferred location in visual space that drives a given voxel’s response (x,y), and 2) amplitude of the voxel’s response (positive or negative) (Fig. 1A). Both positive and negative pRF amplitude was reliable in the DN for all subjects (all z-scored Dice coefficients > 2.9, Fig. 1D). In the dATN, positive amplitude pRFs were reliable in all participants, and negative amplitude pRFs (which constituted a small proportion of the overall pRFs in this network) were reliable in 5/7 participants. For the remainder of the paper, we only consider positive pRFs in the dATN. Importantly, pRF center position was highly reproducible across runs of pRF data in the dATN and DN in all subjects (Fig. 1D).

Next, we compared the retinotopic properties of voxels within the DN and dATN(11, 12, 18). Across participants, we observed a larger proportion of retinotopic voxels in the dATN as compared to the DN (43% of dATN voxels, 9% of DN voxels) consistent with their role as externally- and internally-oriented networks, respectively. Further, the amplitude of retinotopic responses (i.e. whether a stimulus in their preferred visual field location evoked a positive or a negative BOLD response) differed significantly between these two networks (Fig 1E). Consistent with prior reports, negative voxel-wise response amplitudes were clustered within the DN and in similar locations across participants (31, 34), evidenced by a consensus map of pRF sign (Fig. S3). On average, more than half (59%) of all pRFs in the DNs were inverted (i.e., had a negative BOLD response to visual stimulation in their population receptive field, -pRFs). In contrast, just 8% of the pRFs in the dATN were -pRFs (i.e., the vast majority of the voxels had positive BOLD responses to visual stimulation in their receptive field, +pRFs). The difference in negative pRF proportion between the dATNs and DNs was significant (t(6)=7.94, p<0.001; Fig. 1E, inset), and there was no difference among the sub-networks of each system (dATNs: t(6)=0.24, p=0.82; DNs: t(6)=0.89, p=0.41). This distinction is particularly remarkable given the proximity of dATN and DN clusters in posterior cerebral cortex. Like visual areas, both DN and dATN pRFs exhibited a contralateral visual field bias, as we would predict from prior literature (31, 33, 34) (positive dATN pRFs: t(6)=26.0, p<0.001; negative DN pRFs: t(6)=4.86, p=0.003; positive DN pRFs: t(6)=5.20, p=0.002) (Fig. 1F; see Fig. S4 for additional retinotopic properties). Together, these results show that individualized DNs have voxel-wise responses that exhibit visual coding -- the necessary foundation for assessing whether the voxel-wise retinotopic code structures global interactions between the internally- and externally-oriented networks.

Visual responsiveness stratifies DN-dATN resting-state interactions

Next, we examined whether the retinotopic response properties of individual voxels (estimated during the retinotopic mapping sessions) predict their functional coupling across DN-dATN during task-absent, spontaneous activity (during the resting state scans). Specifically, we tested whether voxel-level visual response classes—non-retinotopic voxels, positively tuned pRFs (+pRFs), and negatively tuned pRFs (–pRFs)—predict the strength and direction of coupling between the DN and dATN. Such a finding would reveal a stable, voxel-level organizational principle underlying the global dissociation that is hypothesized to exist between the DN and dATN(11–13, 15, 24, 25).

Across resting state runs, global activity of the DN and dATN was independent (i.e., not correlated) after accounting for variance across all cortical networks using partial correlation(44) (mean correlation = 0. 0.031±0.13 s.d., t(6) = 0.6287, p=0.55; Fig. 2A). Crucially, however, this global relationship masked a striking stratification based on DN voxel responsiveness (Fig. 2B). Non-retinotopic DN voxels (i.e., pRF model R2<0.14) were not significantly correlated with the dATN (mean correlation = 0.014±0.135, t(6) = 0.26, p = 0.79). In contrast, DN voxels that responded positively to visual stimulation (DN positive pRFs, +pRFs) had a positive correlation with the dATN (mean correlation = 0.284±0.152, t(6) = 4.96, p = 0.0025), while DN voxels with systematic negative responses to visual stimulation (DN negative pRFs, -pRFs) were anti-correlated with the dATN (mean correlation = -0.21±0.149, t(6) = -3.75, p = 0.0094). This relationship was further strengthened by adopting more conservative R2 thresholds up to 0.30 to define the pRF population, despite the overall number of included voxels decreasing, suggesting that this effect is not driven by false-positive voxels at the edge of our threshold criteria (Fig. S5).

Retinotopic coding organizes spontaneous interaction between internally and externally oriented brain networks.

A. Overall activity of the dorsal attention networks (dATN-A/B) and default networks (DN-A/B) were independent at rest. One individual’s dATN and DN are shown. Time series from all voxels in the DN and dATN subnetworks were averaged together, and we calculated the partial correlation(44) of these networks while accounting for variance of the other 11 networks within the group-prior parcellation(1, 8). The dATN and DN network correlation was not significantly different from zero, suggesting the activity of these networks is largely independent. Bar plot shows the average partial correlation across participants, each data point is the average of all runs for a single participant. B. Interaction between the DN and dATN differs by visual field preference of DN voxels. Splitting the DN voxels by visual responsiveness revealed distinct patterns of resting-state correlation. DN +pRFs had a positive correlation with the dATN (mean correlation = 0.284±0.152, t(6) = 4.96, p = 0.0025), while non-retinotopic DN voxels (i.e., pRF model R2 < 0.14) were not significantly correlated with the dATN (mean correlation = 0.014±0.135, t(6) = 0.26, p = 0.79). On the other hand, DN -pRFs were anti-correlated with the dATN (mean correlation = - 0.21±0.149, t(6) = -3.75, p = 0.0094). C. Determining spatially-matched pRFs in dATNs and DNs. We assessed the influence of retinotopic coding on the interaction between internally- and externally-oriented brain areas’ spontaneous activity during resting-state fMRI, by comparing the correlation in activation between pRFs in these networks that represent similar (vs. different) regions of visual space. For each DN -pRF, we established the top 10 closest positive dATN pRFs’ centers based on their x and y position (“matched”), and correlated the DN pRF time series with the average timeseries of these matched dATN pRFs. We compared these correlation values with the correlation of the DN pRF timeseries paired with 10 randomly selected pRFs from the 33% furthest (1000 iterations per pRF). We computed this for all DN pRFs for each all runs in each subject. Inset shows the pairwise distances between all pRFs in all participants, dotted line shows the 66%ile. D. Negative (upper) and positive (lower) DN pRFs show retinotopic specific interaction with the dATN. Each set of bars depicts the average correlation of the matched and randomly paired pRFs between the DN and dATN for each participant, with each run shown as connected points. All participants showed greater correlation (either positive or negative) for matched compared with randomly paired pRFs. E. Data shown averaging runs within each participant. For both positive and negative pRFs, the correlation between matched compared with randomly matched pRFs was significantly greater (consistent with their amplitude, ps<0.01). F, G. show the correlation between positive and negative pRFs in the DN with the dATN, split across the sub-networks of the DN-A and B. Both DN-A and B show a significant impact of matching, indicating that both subnetworks interact with the dATN via a retinotopic code (ps<0.05). *: p<0.05, **: p<0.01, ***:p<0.001

This result shows that voxel-level visual response profiles shape DN-dATN coupling during spontaneous resting-state activity. Specifically, the DN and dATN activation is independent during rest. However, at the voxel-level, specific sub-groups of DN voxels have distinct coupling patterns with the dATN that depends on the valence of voxels’ visual responses. DN and dATN voxels with positive visual responses show a positive relationship during rest, and a notable subset of DN voxels with negative visual responses display the canonical opponency with dATN voxels. This suggests that retinotopic coding may be a mechanism that enables visual information to be exchanged between these large-scale brain systems.

Shared visual field preferences structure functional coupling among visually-responsive DN and dATN voxels

We next asked whether functional coupling between DN and dATN voxels is modulated by the visual field preferences of retinotopic voxels — that is, whether their pRFs are centered on similar versus different visual field locations (i.e., their retinotopic alignment). We hypothesized that DN–dATN interactions would be stronger for voxel pairs with shared visual field preferences. To test this, we quantified the retinotopic alignment of voxel-wise pRFs across the DN and dATN.

Because we only considered positive pRFs in the dATN, hereafter we refer to these as “dATN pRFs.”

For each pRF in the DN (+pRFs and –pRFs), we computed the Euclidean distance between its receptive-field center (x, y) and those of all dATN pRFs (Fig. 2C). We then identified the 10 closest dATN pRFs (“Matched” set) and averaged their resting-state time series. As a comparison group, we repeated this procedure using 10 randomly selected, poorly matched dATN pRFs (“Random” set; see Methods). This procedure was repeated 1000 times for each DN voxel across resting-state runs.

Matched dATN pRFs were distributed throughout the network, spanning posterior retinotopic cortex to anterior regions, including prefrontal cortex (Fig. S6). Crucially, the spatial dispersion (i.e., the distance between the pRFs across the cortex) of matched and randomly-matched pRFs did not significantly differ (matched versus random – DN -pRFs x dATN pRFs: t(6)=0.38, p=0.71; DN +pRFs x dATN pRFs: t(6)=0.67, p=0.52), which suggests that spatial proximity is not a confounding factor.

If DN-dATN interactions are scaffolded by a retinotopic code, spatially matched pRFs should exhibit stronger functional coupling than randomly-matched pRFs. Consistent with this prediction, after partialling out variance from DN +pRFs to isolate the specific co-fluctuation between DN - pRFs with dATN pRFs (34, 44), DN -pRFs remained anticorrelated with dATN pRFs in nearly every resting-state run (Fig. 2D). Critically, matched pRFs were significantly more anticorrelated compared with randomly-matched pRFs (linear mixed-effects model with random intercepts for subject and run: F(1,194) = 105.7, p < 0.001; coefficient = –0.023 ± 0.002 s.d.). This effect was evident when run-level correlations were averaged within each participant (7/7 participants rmatched < rrandom; t(6)=6.9, p<0.001, Fig.2E) and could not be explained by signal quality differences, as matched and random pRF sets did not differ in tSNR (t(6) = 1.01, p = 0.353). Finally, to rule out any possible influences from vascular stealing (i.e. the shunting of blood into active tissue from nearby regions), we repeated the matching analysis after excluding any voxels within a 3mm radius of a major vessel (Fig. S7; see Methods). Both matching effects remained after excluding vascularly susceptible voxels (+DN x dATN: t(6) = 6.054; p < 0.001; -DN x dATN: t(6) = -5.0448; p < 0.01). This is strong evidence that the retinotopic code underpins the opponent interaction between negative DN and positive dATN pRFs.

Voxel-wise retinotopic alignment scaffolded the interaction between positive DN and dATN pRFs as well. After controlling for DN -pRFs, DN +pRFs were positively correlated with dATN pRFs (Fig. 2D). Retinotopic alignment again strengthened the interaction: matched correlations exceeded random correlations in a mixed-effects model (F(1,194)= 406.5, p < 0.001; coefficient = +0.050 ± 0.003 s.d.) and for every individual participant (7 / 7; t(6) = 5.7, p = 0.001) (Fig. 2E).

Together, these results demonstrate that shared retinotopic alignment enhances functional coupling between DN and dATN voxels. Voxels tuned to the same region of visual space show stronger spontaneous co-fluctuation—whether positively or negatively coupled—than those with non-overlapping spatial preferences. This suggests that a shared spatial code helps organize spontaneous interactions between large-scale internal and external attention networks during rest.

Subnetwork Specificity within the DN

Recent work has suggested that the subnetworks of the DN, DN-A and DN-B, support different cognitive processes: episodic projection and theory of mind, respectively (8, 45–47). Indeed, DN-A is particularly sensitive to the process of scene construction(47), closely linked to visual imagery, suggesting that dATN pRFs would show greater interaction with visually-responsive voxels in DN-A compared with DN-B. Consistent with this hypothesis, spontaneous correlations with dATN pRFs were significantly stronger for DN-A than DN-B voxels (collapsed across positive and negative pRFs: t(6) = 4.80, p = 0.003). Crucially, however, voxel-scale retinotopic alignment structured interactions between the DN and the dATN for both DN sub-networks (Fig. 2F-G; DN-A: -pRFs t(6) = –4.56, p = 0.004; +pRFs t(6) = 5.65, p = 0.0013; DN-B: -pRFs t(6) = –7.49, p < 0.001; +pRFs t(6) = 2.92, p = 0.027). The strength of this effect did not significantly differ between DN-A and DN-B (t(6) = 1.51, p = 0.182). Thus, although DN-A is more strongly globally coupled with dATN, retinotopic preferences of individual voxels structure functional interactions across both DN subnetworks equally.

In addition, using the same matched-vs-random pRF procedure, we found that resting activity between the lateral place-memory area (LPMA) and the adjacent occipital place area (OPA) was significantly more correlated for retinotopically matched than random voxel pairs (t(6)=3.01, p=0.024; SI). This shows that retinotopic scaffolding within domain-specific systems extends beyond perceptual and memory tasks (34), organizing spontaneous activity within these systems as well. These findings build on and extend our previous work (34) and others(48).

Retinotopic coding is intrinsic to internally-oriented areas

Our findings thus far highlight a role for retinotopic coding in structuring spontaneous interactions between a subset of voxels across large-scale internally- and externally-oriented brain networks. However, a key mechanistic question remains: is retinotopic coding an intrinsic property of the DN, or does it only emerge in response to input from sensory systems and lower-level networks like the dATN? One possibility is that DN-driven modulation of dATN activity reflects a global, non-selective suppression, with retinotopic coding in the DN arising only through input from the dATN. Alternatively, retinotopic coding may be intrinsic to the DN itself – such that signals projecting from the DN to the dATN are retinotopically-specific, allowing the DN to shape fine-grained spatial representations within the dATN. Disambiguating these possibilities is essential for determining whether the retinotopic scaffold constitutes a general mechanism for cross-network communication, independent of task demands or cognitive domain (e.g., memory versus perception).

To address this, we again turned to participants’ resting-state fMRI data. Because participants are not engaged in any task, resting-state provides a neutral testbed for assessing spontaneous DN–dATN interactions. Moreover, experience sampling during rest has shown that attention naturally alternates between internal and external focus during rest (49, 50), and thus is likely to result in activation of both the DN and dATN, making it possible to assess whether retinotopic coding emerges primarily during periods of elevated DN or dATN activity.

We developed an event-triggered analytical approach to dissociate periods when activity of the retinotopic voxels within the dATN and DN was elevated. We refer to these periods when dATN or DN voxels’ activity was elevated as dATN-driven from DN-driven events, respectively. Specifically, we identified timepoints where spontaneous activity of individual DN or dATN pRFs exceeded the 99th percentile of their resting-state distribution. These high-activation time points were treated as neural “events” (Fig. 3A-C). To test whether these events produced retinotopically-specific responses in the target network, we compared peri-event activation in the target network’s most 10 closely matched pRFs (i.e., those with similar visual field preferences) to that in the 10 worst-matched (anti-matched) pRFs (Fig. 3A). We used the worst matched pRFs for this analysis because the large number of pRFs made the random matching procedure impractical. All analyses were conducted separately for positive and negative pRF populations within the DN. This design enabled us to test two competing predictions: if retinotopy in the DN only reflects input from more canonically perceptual networks like the dATN, we should only see retinotopically specific coupling during dATN-driven events. If, however, the DN possesses an intrinsic retinotopic scaffold, we should observe retinotopic specificity during both dATN- and DN-driven events. A windowed correlation analysis produced similar results (Fig. S8).

DN-driven vs. dATN-driven neural events detected in spontaneous resting-state dynamics show evidence for retinotopically-specific suppression.

A. Event detection and analysis procedure and example events from a single resting-state run. To detect events, we extracted the time series from each pRF in the source regions (positive or negative pRFs in the DN; positive pRFs in the dATN) and isolated time points where the z-scored time series exceeded 2.4 s.d. (99th percentile). We then examined the activity of 10 best matched and 10 worst matched (anti-matched) pRFs from the target region in this peri-event time frame (6 TRs (8 s) before and after the event). B. Example DN-driven event in the representative time series shows elevated source pRF activity (DN -pRF) and corresponding target area activity (dATN +pRF). C. Same as B, but for dATN-driven activity with the event detected in a dATN +pRF. D, E. Evidence for retinotopically-specific suppression of ongoing activity during DN-driven events detected in DN -pRFs (D) and dATN-driven events detected in dATN +pRFs (E). D. Time series show the peri-event activity of matched and anti-matched dATN +pRFs averaged across all participants for events detected in DN -pRFs (N=78,978 events). E. Time series shows the peri-event activity in DN -pRFs during events detected in dATN +pRFs (N=541,546). F. For DN -pRFs paired with dATN +pRFs, target area shows retinotopically-specific suppression of activity for both DN-driven and dATN-driven events. Bars show the average activation at event time of the target areas’ matched and anti-matched pRFs for each participant. Activity was lower in matched compared with anti-matched pRFs for DN-driven and dATN-driven events (ts>3.8, ps<0.009). G, H. Evidence for retinotopically-specific excitation of ongoing activity during events DN-driven events detected in DN +pRFs (G) and dATN-driven events detected in dATN +pRFs (H). G. Grand average time series for matched and anti-matched dATN +pRFs during events detected in DN +pRFs for all participants. H. Grand average time series for matched and anti-matched DN +pRFs during events detected in dATN +pRFs. I. For DN +pRFs paired with dATN +pRFs, target area shows retinotopically-specific excitation for both DN-driven and dATN-driven events. Bars show the average activation at event time of the target areas’ matched and anti-matched pRFs for each participant. Activity was higher in matched compared with anti-matched pRFs for DN-driven and dATN-driven events (ts>8.75, ps<0.001). Interestingly, the activity of anti-matched pRFs was greater during DN-driven compared with dATN-driven events (t(6)=4.92, p=0.0026), suggesting that input from DN +pRFs to the dATN +pRFs may be more global and less retinotopically-targeted compared with dATN-driven signaling between these networks. For time series plots, solid bars along the x-axis indicate periods with a significant difference between matched and anti-matched pRF activity within the target area. *: p<0.05, **: p<0.01, ***:p<0.001. -DN: Negative DN pRFs; +DN: positive DN pRFs; +dATN: Positive dATN pRFs

Across all participants, we detected 78,978 events in DN -pRFs, 52,595 events in DN +pRFs, and 541,546 events in dATN pRFs. dATN-driven events were more frequent than either DN-driven type (dATN vs. negative DN: t(6)=15.19, p<0.001; dATN vs. positive DN: t(6)=25.63, p<0.001), but there was no difference between DN-driven event types (t(6)=2.03, p=0.18). On average, participants exhibited 2082±535 (mean±SD) negative DN, 1799±400 positive DN, and 3102±602 dATN events per run. Each pRF showed 0–4 events per run (Fig. S9). More than 90% of timepoints involved <10 simultaneous events, suggesting that individual pRF-level correlation could be isolated from global fluctuations in regional activity, making this approach suitable for evaluating distinct interactions at the individual voxel level.

We found a robust effect of retinotopic alignment (i.e., matched versus anti-matched) on pRF activation during both dATN- and DN-driven interactions (Fig. 3D-I). For DN -pRFs, a linear mixed-effects model (factors: DN-driven/dATN-driven and retinotopic alignment; subject as random effect) revealed a main effect of match status (F(1,24)=15.3, p=0.0006) and no interaction between match status and direction (F(1,24)=0.0748, p=0.40). During DN-driven events in DN - pRFs, matched dATN pRFs showed lower activity than anti-matched (t(6)=5.46, p=0.001) and were significantly deactivated relative to the pre-event baseline. The same was true of DN -pRFs during dATN-driven events: we observed lower activation for matched vs. anti-matched pRFs (t(6)=3.80, p=0.009). These results are consistent with a retinotopically-specific interaction between dATN +pRFs and DN -pRFs during both DN-driven and dATN-driven events detected in spontaneous activity during resting-state fMRI.

The interaction between DN +pRFs and dATN pRFs also showed evidence of retinotopic coding during both dATN- and DN-driven events, but these effects were modulated by a direction × match interaction (F(1,24)=12.84, p=0.002). Both DN-driven (t(6) = 8.76, p < 0.001) and dATN-driven (t(6) = 9.14, p < 0.001) events evoked strong activation in the target network (dATN or DN, respectively). However, anti-matched pRFs were more active during DN-driven compared to dATN-driven events (t(6) = 4.92, p = 0.0026), suggesting that signals from positive DN pRFs to dATN may be less retinotopically specific than incoming signals from the dATN. These results align with prior findings showing decreased spatial precision in visual representations during memory recall compared to perception(51, 52).

Taken together, these results support the view that retinotopic coding serves as a general scaffold for DN-dATN interactions. This alignment shapes spontaneous coupling during both periods of DN-driven and dATN-driven activity, even in the absence of external tasks. Because the effect persists without sensory input, it likely reflects an intrinsic feature of the DN activity rather than a transient result of drive from sensory input.

Discussion

Despite extensive evidence that internally- and externally-oriented brain networks – including the default network (DN) and dorsal attention network (dATN) – are critical for human cognition(1–8, 11–13), the coding principles that enable these distributed systems to interact have remained unclear(30, 53–57). Here, we show that DN–dATN coupling is retinotopically grounded at the voxel scale: the visual field biases of individual voxels scaffold bidirectional interactions between these networks, even during spontaneous dynamics in the absence of visual task demands. Analyses of transient elevated activity in either the DN or dATN confirm that this retinotopic opponency shapes both dATN-driven signals to the DN and DN-driven signals to the dATN, indicating that the retinotopic code is intrinsic to both internal and external networks. These findings offer a multi-scale account of neural communication, in which interactions among sub-populations of voxels with shared tuning preferences are nested within macro-scale network dynamics. Nesting multiple neural codes might enable ongoing computations within a larger brain system (e.g., attending to internal mental states within the DN during memory recall), while simultaneously allowing for the sharing of fine-grained representations across brain systems (34). Here we show this latent, voxel-scale interaction persists even without structured behavioral demands, opening the opportunity to study these fine-scale interactions across brain networks in resting-state data (58).

Resolving the coding principles of the DN is critical given its central roles in spatial and episodic memory, social cognition, and executive function(13, 59–64). Classic views of systems-level brain organization posit regions at the apex of the cortical hierarchy, like the DN, likely have abstract, amodal neural codes, in contrast to sensory codes–like retinotopy–that manifest in earlier areas of the cortical hierarchy (2, 29, 30). Our findings and others challenge this view. Instead of shedding the retinotopic code at the cortical apex, the DN retains a retinotopic code(31, 33) and functionally deploys this code to structure interactions with the externally-oriented dATN. Thus, a code once thought to be exclusive to early visual cortex and visually-oriented tasks plays a widespread role in shaping brain activity across the cortical hierarchy—even at rest.

Establishing this voxel-level structure within the DN and dATN reframes a long-standing debate that has cast these networks as globally competitive or functionally independent(11, 23, 27, 28).

These views have stemmed from macro-scale observations of reciprocal activation patterns during external versus internal task demands(6, 10, 13, 14, 16–23), and resting-state analyses that reveal zero or negative correlations between their mean time series(11, 12, 24, 25). However, by focusing on global, network-level activity across voxels, these approaches have obscured a latent, voxel-level structure to DN-dATN functional coupling. Specifically, we find that DN–dATN coupling is not independent. Instead, two subpopulations of voxels within the DN contribute to a retinotopically-grounded cross-network interaction that is stratified by their voxel-level visual selectivity. One population of DN voxels is positively tuned to visual stimulation, and positively covaries with dATN voxels that share their visual field preference. In contrast, DN voxels that are negatively tuned to visual stimulation negatively covary with retinotopically matched dATN voxels as a function visual field preference. When these voxel-level tuning preferences are not taken into account, global cross-network opponency or independence(11, 12) reinforces the misconception that the networks support mutually exclusive cognitive modes. By revealing a shared retinotopic scaffold, our findings reconcile why DN and dATN can appear independent in coarse analyses yet remain functionally integrated, offering a new framework for understanding how internal and external attention are coordinated in the human brain.

Importantly, there are key differences in the nature of retinotopic coding between the DN and visual areas, including the dATN. First, retinotopy in the DN is predominantly negative in valence—DN voxels more often show spatially-specific suppression in response to visual stimulation(31, 33, 34), unlike the positive spatially-specifical response that is typically observed in visual areas. Second, the DN lacks the map-like retinotopic organization seen in early visual cortex. In visual areas, retinotopic coding is not only the foundational code, but also the dominant topographic organizing principle across the cortical surface, where voxels that are proximal on the cortical surface tend to share visual selectivity for nearby regions of visual space(65). Prior work has emphasized the visual response bias in regions where voxel-wise retinotopic responses lack a map-like organization(35); overall, the DN does exhibit this kind of bias. However, our results show that the voxel-scale activity underpinning this bias reflects the latent connectivity of those voxels. Thus, we adopt the term “retinotopic coding”, because this voxel-scale coding scheme exists without a map-like organization on the cortical surface. For example, rodent and bat head direction cells are not laid out in a literal ring, but the population code of these neurons forms a ring manifold(66, 67). Similarly, in humans, retrosplenial cortex shows head-direction coding during navigation tasks despite lacking topographic organization of head-direction responses(68). Analogously, the DN may implement a retinotopic code that is distributed across the network in a format that does not spatially mirror the retina. That is, the DN may use a fine-scale retinotopic code without a retinotopic map – the representational structure can be preserved even in the absence of spatially ordered cortical layouts.

What function might the bivalent (positive and negative) voxel-scale retinotopic alignment across these large-scale networks serve? Our previous work has identified a role for retinotopic-alignment in structuring the functional coupling of scene-selective mnemonic and perceptual brain areas(34, 48) located near the dATN and DN, respectively (36, 69). Near to the DN, scene-selective memory areas exhibit a negative retinotopic code, while functionally-paired scene-selective perceptual regions exhibit a positive retinotopic code(34). This bivalent retinotopic code structures the mutual activity of these mnemonic and perceptual areas: scene-recall activates scene-memory voxels, it but suppresses voxels in scene-perception areas that are tuned to matched visual locations, a pattern that reverses during scene perception(34). The current work significantly extends this observation beyond perceptual and mnemonic task-states -- identifying widespread retinotopic alignment across large-scale networks with presumed different orientations (internal vs. external) even during spontaneous activity at rest, in the absence of tasks.

As to the functional role of the bivalent retinotopic alignment across internal and externally-oriented large-scale networks, we propose two complementary possibilities. One possibility is that the bivalent (positive and negative) subpopulations of DN pRFs may serve as a neural mechanism for maintaining separate channels for perceptual and mnemonic information(54), preventing mnemonic signals from being interpreted as or interfering with sensory signals. A second possibility is that bivalent coding across the DN and dATN could be a mechanism for signaling mnemonic predictions via top-down inhibitory processes(34, 70). Specifically, when the DN generates predictions about incoming sensory information based on memory, the inverted retinotopic code could act as a suppressive signal that sharpens tuning in perceptual areas and enhances signal-to-noise for expected stimuli(71, 72). These mechanisms may work in concert: the inverted code simultaneously maintains separate channels for memory/perception within the DN, while enabling precise top-down predictive signaling across DN-dATN. Future work that combines high-resolution recordings with behavior will be essential for testing how these mechanisms jointly support constructive operations such as memory-based prediction during visually guided navigation.

Relatedly, the DN is considered a transmodal hub for cortical processing, where disparate sensory and motor processes converge (59, 73) The DN’s position at the cortical apex implies connections with and influence over unimodal cortical areas. However, the mechanism linking unimodal and transmodal networks had been unknown. Prior work posited that sensory coding in transmodal areas might serve this function (31, 35), and our data provide direct empirical support for this account: specific visually-responsive voxels provide an input/output interface bridging perceptual and memory systems. This complements work delineating specific affective and effective subregions within the DN that link the DN to other brain areas (74). Thus, while the DN may be “distant from input”(28), these data suggest that it is not disengaged from sensory processing.

Retinotopic alignment between the DN and dATN was clear in both subnetworks of the DN, DN-A and DN-B(3, 8, 20). This is surprising given their distinct functional profiles: DN-A has been more directly linked to visually grounded memory processes, such as episodic projection and scene construction(8, 20, 47), and spatially overlaps with regions previously shown to exhibit retinotopy and engage in scene memory and perception(69). In contrast, DN-B is more often associated with processes that are not obviously tied to vision: theory of mind and affective tasks, such as interpreting false beliefs or representing physical and emotional pain (8, 20). Why might the DN-B show retinotopically-grounded coupling with perceptual networks? Retinotopic coding might still be a useful format for transforming abstract social representations into sensory-grounded signals to inform perceptual predictions (e.g., anticipating changes in facial expression) or behavioral decisions (e.g., making eye movements). Future work might consider direct manipulation of social perceptual processes to assess the impact of social tasks on retinotopic coding between DN-B and dATN.

Moments of high spontaneous activity originating in the DN co-occurred with suppression of retinotopically-matched dATN pRFs, suggesting that retinotopy functions as a native coding format in the DN for targeted modulation of perceptual cortex. This result aligns with other recent studies suggesting an important role of the DN in shaping visual processing. For example, ongoing pre-stimulus DN activity has been shown to influence sensitivity during near-threshold visual object recognition(75). Other work has shown that subareas of the DN encode kinematic of spatial exploration in rodents(76), and represent the visuospatial context that is associated in memory with a perceived scene(34, 36, 69, 77, 78). Relatedly, DN activity reflects semantic-level attentional priorities during visual search(79). Taken together, these findings show that, rather than being disengaged during visual tasks, the DN actively shapes responses in perceptually-oriented cortex in a spatially-precise manner. In this sense, retinotopic coding is part of the DN’s “native language”. Taken together, these findings prompt a reevaluation of the DN’s role in perceptual processing, and the extent to which roles in “internally- and externally-oriented attention” adequately captures the DN and dATN’s role in cognition.

Is the DN a “visual” network? One possible interpretation of retinotopic coding in the DN is that the DN directly and obligatorily represents retinal position in addition to more abstract features(38, 65). Under this hypothesis, the retinotopic code may indicate that the information represented in the DN is, in part, sensory. Alternatively, the apparent visual coding may simply represent an underlying connectivity structure(80), which enables associative and semantic information that the DN directly represents(30, 81, 82) to be effectively transferred to sensory areas of the brain. Under this alternative hypothesis, the information represented in the DN is not itself sensory in nature; the retinotopic code in the DN simply serves as a “highway” between the abstract representations in the DN and sensory representations in the dATN. Notably, in either case the retinotopic scaffolding represents a latent structure linking high-level and low-level areas through consistent, spatially-organized interactions(80). This provides more evidence that low-level sensory codes are distributed across the brain, joining recent work showing latent somatotopic codes in regions of the brain that are not typically associated with somatosensation, including visual cortex(83). The wide and overlapping distribution of sensory codes across the brain may allow for multiplexing of complex information to structure cross-network interactions(35, 83). Further studies investigating this framework with diverse tasks and stimuli could reshape our understanding of how the brain processes and integrates information across different levels of cognitive complexity.

Our study and other recent studies of voxel-wise visual responses in traditionally non-visual areas (31, 33, 34, 48, 84–86) raise important questions about model selection and thresholding criteria. In canonical visual cortex, pRF analyses often rely on variance-explained thresholds to identify reliable voxel-wise model fits. However, it is less clear what R² criterion should be applied in non-visual regions, where responses may be weaker or more heterogeneous. To address this issue, we designed a method to determine a threshold for voxel-wise model fits (R2) empirically by leveraging resting-state fMRI data as a noise model. Crucially, pRF parameter estimates (center position and amplitude) for voxels that passed this empirical threshold were reliable across runs, suggesting this method robustly selects voxels with stable pRF properties. This empirical method may be used evaluate retinotopic responses across the brain in future studies (86).

Our study made use of the NSD, which included pRF modeling data to characterize visually-responsive cortex in a small group of individual participants. One limitation of this dataset is the small number of pRF task runs, which might lead to noisier model estimates. Our robust thresholding approach mitigates this concern. Furthermore, our primary results were based on voxel-wise pRF alignment within independently acquired resting-state fMRI data; noisy estimates of pRF parameters would obscure this relationship and make our results less likely. Relatedly, here we used a single gaussian model, consistent with prior work on negative visual responses in memory systems (31, 33, 34). However, other models of visual receptive fields might offer further insight into the DN’s visual responsiveness, such as double gaussian models of surround suppression (87) or compressive summation (88). Future studies might directly compare different visual models to further refine the computations underpinning visual responses in the DN.

In summary, our results show that a voxel-level retinotopic code organizes the spontaneous interactions between large-scale internally- and externally-oriented networks in the human brain – the DN and dATN. Even when global activity appears uncorrelated across DN-dATN, voxel-level visual-field tuning reveals a tightly interlocked, bidirectional functional dynamic that persists in the absence of external stimulation. This latent scaffold reconciles the apparent macroscale independence of internally- and externally- oriented systems with their evident functional need to cooperate, revealing that the DN is not divorced from perceptual networks, but modulates perceptual voxels through a common visuospatial reference frame. Taken together, by grounding large-scale network interactions in a shared, bivalent spatial code, retinotopy may offer a simple organizing principle for the brain’s coordination of thought and perception.

Materials and Methods

Dataset and Participants

The data analyzed here are part of the Natural Scenes Dataset (NSD), a large-scale precision 7T fMRI dataset consisting of high-resolution anatomical scans, retinotopic mapping, functional localizers for visual areas, and both task and resting-state fMRI data for eight adult participants. A full description of the dataset can be found in the original manuscript (37). Here, we detail the data, processing, and analysis steps relevant to the present work.

Participants (N = 8; 2 male, 6 female; ages 19–32) were scanned at the University of Minnesota over the course of 33-43 sessions. All participants had normal or correct to normal vision and no known neurological impairments. Informed consent was collected from all participants, and the study was approved by the University of Minnesota institutional review board. One subject (subj03) was excluded from resting-state analyses for having insufficient resting-state runs that passed our quality metrics (see below).

For the present study, we focused on two components of the NSD protocol: retinotopic mapping and resting-state fMRI. All participants completed three runs of a sweeping bar retinotopic mapping task (5 mins each). Retinotopic mapping was conducted during Session 1, along with the functional localizer scans. This task is referred to as “RETBAR” in NSD documentation.

All participants completed at least 14 runs of a resting-state fixation task (range: 14-34 runs; 5 mins each). Resting-state data runs were included in all NSD scan sessions after session 20 (i.e., session 21, 22, etc.). Each of these sessions included two 5-minute resting-state runs -- one at the start and one at the end of the session -- yielding 14 to 34 total resting-state runs per participant. During the second resting-state run, participants performed a deep breath in the first 5 seconds of scanning. To account for this, and to ensure the two resting-state runs were equated in length, we excluded the first 30 seconds of the resting-state data from all runs. One participant (subj03) was excluded from resting-state analyses due to having an insufficient number of usable runs after quality control. All remaining participants had at least 8 high-quality resting-state runs (mean: 14 ± 6.13 runs).

MRI acquisition and processing

For this study we made use of the following data from the NSD: anatomical (T1 and FreeSurfer segmentation/reconstruction(89, 90)), functional regions of interest (ROIs), and minimally preprocessed retinotopy and resting-state time series. All analyses except network identification were conducted in original subject volume space, and data were projected on the surface for visualization purposes only. Network identification was performed on the fsaverage6 surface, then projected into subject specific volume space for quantification.

Anatomical data

Anatomical data was collected using a 3T Siemens Prisma scanner and a 32-channel head coil. We used the T1-weighted anatomical images provided at 0.8mm resolution as well as the registered output from FreeSurfer recon-all, aligned to the 1.8mm functional data. For venography, we used the provided time of flight (TOF) MR venography images with a resolution of 0.5mm3. For visualization purposes, we projected statistical and retinotopy data to the cortical surface using SUMA (91) from the AFNI software package(92).

Functional MRI data acquisition and processing

All analyses were conducted on 1.8mm isotropic resolution minimally processed runs of a sweeping-bar retinotopy task and resting-state fixation provided in the NSD.

Quality assessment

To ensure only high-quality resting-state data were included, we trimmed the first 25 TRs (40s) from each run (77), which left 4.25 min of resting-state data per run. We then used AFNI’s quality control assessment tool (APQC(93)) on the raw trimmed resting-state time series to assess the degree of motion in the resting-state scans. We excluded runs with greater than 1.8mm maximum displacement per run or 0.12mm framewise displacement from analysis, and assessed runs with greater than 1mm displacement or 0.10mm framewise displacement on a case-by-case basis(8). After exclusions, one participant (subj03) had only a single run that survived our criteria (4.25 minutes of data)), so we excluded them from our analyses. The remaining 7 participants had at least 8 resting-state runs (> 34 minutes of data; mean number of runs: 14±6.13 (sd), range: 8-24 runs).

ICA denoising

To further denoise the retinotopy and remaining resting-state data, we used manual ICA classification of signal and noise on the minimally-preprocessed time series(36, 40, 41). We used manual classification because automated tools perform poorly on the high spatial and temporal resolution data of the NSD(94). For each retinotopy and resting state run, we decomposed the data into independent spatial components and their associated temporal signals using ICA (FSL’s melodic(95, 96)). We then manually classified each component as signal or noise using the criteria established in(41). Noise signals were projected out of the data using fsl_regfilt(40).

We did not perform global signal regression(26). Data were normalized to percent signal change. A 2.5mm FWHM smooth was applied on the surface to resting-state fixation data used in functional connectivity network identification. The analysis of voxel-scale interactions performed on unsmoothed data.

Data analysis – Identifying Resting-State Networks

Individual-specific cerebral network estimation

In each participant, we identified a set of 15 distributed networks on the cerebral cortical surface using a multi-session hierarchical Bayesian model (MS-HBM) functional connectivity approach (8, 39). Briefly, the MS-HBM approach calculates a connectivity profile for every vertex on the cortical surface derived from the top 10% of correlations from that vertex to 1,175 ROIs uniformly disturbed across the cortical surface(1). We applied an expectation-maximization algorithm, initialized with a 15-network group-level spatial prior created by applying previous methods of network identification(1) on a portion of the HCP S900 data, to all the connectivity profiles from all participants. This algorithm then estimated each network’s group-level connectivity profile, along with parameters for subject-, session/run-, and region-level variability, which together were used to generate individual-specific, labeled parcellations.

The MS-HBM approach is well-suited to derive parcellations for the participants in the NSD, where resting-state data was collected across multiple days, because this method uses repeated resting-state runs for each participant to estimate within-participant variance. This leverages a strength of the NSD (the relatively large number of resting-state runs for each individual) and can account for differences in connectivity both within an individual (likely the result of confounding variables such as scanner variability, time of day, etc.), and between participants (reflecting potentially meaningful individual differences). Moreover, the MS-HBS incorporates a group spatial prior, further stabilizing the resulting network parcellations for each participant. This also ensures that all networks will be identified in all participants, while allowing for idiosyncratic topographic differences.

Data analysis – Retinotopic Mapping

Population receptive field (pRF) Modelling

The NSD retinotopy stimuli consist of a mosaic of faces, houses, and objects superimposed on pink noise that are revealed through a continuously drifting aperture. For our analysis, we considered only the bar stimulus time series, which is consistent with other studies investigating negative pRFs in high-level cortical areas (31, 33, 34). We did not consider the wedge/ring stimulus for any analyses.

After denoising the retinotopy data, we averaged the three retinotopy runs with the bar aperture together to form the final retinotopy time series. We performed population receptive field modeling using AFNI following the procedure described in (43). First, because the pRF stimulus in the NSD is continuous, we resampled the stimulus time series to the fMRI temporal resolution (TR = 1.333s). Next, we implemented AFNI’s pRF mapping procedure (3dNLfim). Given the position of the stimulus in the visual field at every time point, the model estimates the pRF parameters that yield the best fit to the data: pRF amplitude (positive, negative), pRF center location (x, y) and size (diameter of the pRF). Both Simplex and Powell optimization algorithms are used simultaneously to find the best time series/parameter sets (amplitude, x, y, size) by minimizing the least-squares error of the predicted time series with the acquired time series for each voxel. Relevant to the present work, the amplitude measure refers to the signed (positive or negative) degree of linear scaling applied to the pRF model, which reflects the sign of the neural response to visual stimulation of its receptive field.

Identifying Retinotopic Voxels – Empirically-derived R² thresholds

We developed a rigorous, data-driven approach to determine whether our pRF model fits truly reflect visual coding rather than physiological noise. To do this, we leveraged the resting-state fMRI data – where there is no dynamic visual stimulation -- to create a control (null) distribution for retinotopic model fits and establish a “noise floor” for each network (i.e., the DN and dATN). Then, we compared task-derived model fits to this rest-based null distribution to establish an empirical R² threshold for pRFs in each network.

Specifically, we fit the pRF model to the average timeseries across all resting-state runs for each participant. We then compared the resulting voxel-level R² distribution to that from the pRF mapping task data, pooled across participants. To test whether the model fits were significantly higher for the task vs. null distribution, we performed a one-tailed Kolmogorov-Smirnov test for each network on the respective cumulative distribution functions (H0: CDFrest(R2) ≤ CDFtask(R2), Ha: CDFrest(R2) > CDFtask(R2)). These analyses were performed separately for the dATN and positive and negatively responsive voxels within the DN.

We defined a voxel as exhibiting a retinotopic code (hereafter referred to as a “pRF”) if its model fit exceeded 14% variance explained. This threshold was based on the 95th percentile of the null R² distributions, which yielded noise floors of R² = 0.124 (dATN), R² = 0.132 (DN-positive), and R² = 0.137 (DN-negative). We therefore set our global threshold for visually responsive voxels across all networks at R² > 0.14, as this value exceeded the noise floor calculated in each network separately.

Quantifying visual field coverage

Visual-field coverage (VFC) plots represent the sensitivity of an ROI across the visual field. We followed the procedure in Steel et al., 2024(34), which we have reproduced here. Individual participant VFC plots were first derived based pRF center, as well as pRF size and R2. These plots combine the best Gaussian receptive field model for each suprathreshold voxel within each ROI by calculating the max operator at each point in the visual field, defined as the maximum value from all pRFs within the ROI. The resulting coverage plot thus represents the maximum envelope of sensitivity across the visual field. Individual participant VFC plots were averaged across participants to create group-level coverage plots.

To compute the elevation biases, we calculated the mean pRF value (defined as the mean of the average max operator values in a specific portion of the visual-field coverage plot) in the contralateral upper visual field (UVF) and contralateral lower visual field (LVF) and computed the difference (UVF–LVF) for each participant, ROI and amplitude (+/−) separately. A positive value thus represents an upper visual-field bias, whereas a negative value represents a lower visual-field bias.

Reliability of pRF amplitude and center position estimates

We tested whether key pRF features of interest – amplitude (i.e., positive vs. negative) and center position (i.e., x,y) – were reliable across retinotopy runs. To assess the reliability of pRF amplitude, we iteratively compared the amplitude of significant pRFs from individual runs of pRF data. Specifically, for each retinotopy run, we fit the pRF model and then binarized any significant voxels (R² > 0.14) by amplitude. When testing positive-pRF reliability (in dATN and DN), positive voxels = 1; all others = 0. For negative-pRF reliability (in DN), significant negative voxels = 1. We then computed a three-run Dice coefficient: (dice =3∣run1∩run2∩run3∣∣run1|+|run2|+|run3∣). We compared this value against 5000 null Dice scores derived from the same number of voxels randomly sampled from all voxels, both significant and non-significant, in the ROI. For each participant, this resulted in 1 “observed” dice coefficient along with 5000 null values representing the distribution of dice coefficients expected by chance. The observed coefficient was z-scored against this null distribution for each participant.

We also evaluated center position (x,y) reliability. For every significant voxel, we calculated the Euclidean distance between its center estimates across the three run pairs (i.e. 1-2, 2-3, 1-3). As with amplitude, we generated a null distribution by repeating this computation on 5,000 random voxel sets (significant + non-significant). To establish significance of the true estimate, the observed estimate was z-scored against this null distribution for each participant.

Data Analysis – Retinotopically-Grounded Resting-state Analyses

Evaluating global functional coupling between cortical networks at rest

To assess global functional coupling between the default network (DN) and the dorsal attention network (dATN) at rest, we used a partial correlation analysis that accounted for variation associated with other cortical networks (44). For this and all subsequent analyses, each individual resting state run was z-scored in time. Prior to calculating the partial correlation between them, we first averaged the time series within each DN and dATN subnetwork.

Evaluating the Contributions of Voxel-Level Visual Response Valence to Cross-Network Coupling

To test whether DN–dATN coupling is modulated by the visual responsiveness (i.e., response valence) of DN voxels, we repeated this analysis separately using only: (i) sub-threshold/non-retinotopic voxels, (ii) positive pRFs, or (iii) negative pRFs from the DN -- while still partialing out the remaining networks. This allowed us to assess whether specific classes of visually responsive voxels in the DN differentially contribute to functional coupling with positive pRFs in the dATN.

Evaluating Voxel-Level Retinotopic Alignment in Cross-Network Coupling

To test whether functional coupling is shaped by the shared visual field preferences of DN-dATN voxels (i.e., retinotopic alignment), we compared resting-state correlations between pRF pairs across the DN and dATN that had matched (vs. unmatched) pRF centers.

To do this, for each pRF in the DN (positive or negative), we computed the Euclidean distance between its receptive-field center (x, y) and those of all dATN +pRFs. We focused on center position instead of pRF size because 1) size estimates are relatively unstable compared with center position(97), and 2) we lacked a clear prediction about how pRF size would be related between the two networks. For each DN voxel, we identified the 10 nearest dATN +pRFs based on center position—the “matched” set—and averaged their resting-state time series. As a null comparison, we defined a “random” set by sampling with replacement 10 dATN +pRFs from the most distant third of the match distribution (i.e., (the 1/3 furthest pRF pairs). We considered the bottom 33% because the overwhelming majority of pRFs constituted relatively equal “good” matches. This random sampling was repeated 1,000 times per DN voxel, generating a distribution of 1,000 random correlation values, which we then averaged to yield a single “random” correlation for each DN source pRF.

We calculated the correlation between each DN pRF’s resting-state time series and that of its 10 best-matched dATN pRFs (“matched” correlation). We averaged across the 10 best matches to (1) reflect the likely distributed input from multiple dATN sources and (2) improve signal-to-noise. To isolate pRF-specific effects, we partialed out the mean time series of the complementary DN pRF population from the signals of interest. For example, when analyzing correlations between – DN and +dATN, we treated +DN activity as a nuisance regressor (and vice versa)(34, 44). This provided a network-level analog to the whole-brain partial correlation described above (see Section: Correlation among cortical networks), allowing us to isolate the unique contribution of the pRFs of interest from overall activity in the region as well as to control for global fluctuations that cause widespread positive correlations like motion and attentional state(34, 44). All correlation values were Fisher-transformed, and we took the median value across voxels within each network to yield a single value per run. This procedure was repeated for all runs and all participants.

We then statistically compared Fisher-transformed correlations for matched vs. random pRF pairs using a linear mixed effects model, with match status (matched vs. random) as a fixed effect and subject and run as random effects. This model accounted for both between- and within-subject variation across session. For follow-up tests, we compared the average matched and random values across participants using a paired t-test.

Masking Vascularly Sensitive Voxels

To ensure that the above retinotopic alignment analyses were not the result of large-scale vascular features (e.g. vascular stealing), we created whole brain mask to remove any voxels within 3mm of a significant vessel. We identified and enhanced vessels within the 0.5mm TOF MR venography image using a Hessian-based multiscale vesselness filter (Matlab command fibermetric), then binarized the result based on a threshold calculated from a fit 2-Gaussian Mixture (to isolate vessels from background noise). Finaly, we dilated the vessels using a spherical structuring element with a radius of 3mm (6 voxels), and inverted the image to produce a mask excluded vessel-adjacent voxels.

Resting-state event detection analyses

To dissociate the influence of DN drive versus dATN drive on spontaneous interactions, we developed an event-triggered analysis of resting-state fMRI data. Because participants were not engaged in any task, resting-state data provide a neutral testbed for evaluating the directionality and spatial specificity of DN–dATN interactions. Building on prior findings that attention during rest naturally alternates between internally and externally oriented states(23, 49, 50), thought to be supported by the DN and dATN respectively, we reasoned that transient “neural events” in these networks could reveal spontaneous DN-driven and dATN-driven communication.

We defined DN-driven events as transient high-activation periods in DN pRFs, and dATN-driven events as high-activation periods in positive-polarity dATN pRFs(98, 99). Event detection was performed independently for each source pRF by taking each pRF’s z-scored resting-state time series and identifying TRs with unusually high activity. Specifically, we considered any time point with a z-value greater than 2.4 (i.e., 99.18 percentile) as a neural event. Results were comparable with a range of thresholds (2.1 < z < 2.9).

At each detected event, we extracted the resting-state time series of the 10 most retinotopically aligned pRFs in the target network—i.e., the 10 with the smallest Euclidean distance in receptive-field center (x, y)—over a 13-TR peri-event window (–6 to +6 TRs). These time series were normalized to the mean of the first four pre-event TRs to allow comparisons across events and conditions. To evaluate the spatial specificity of these interactions, we compared responses in the matched pRFs to those in an “anti-matched” pRFs set: the 10 most distant pRFs in the target network in terms of visual field preference (i.e., pRF center positions). Unlike prior analyses that used random sampling to define unmatched control pRFs, here we used a deterministic anti-match procedure to avoid the computational burden of repeated resampling, given the large number of events.

This procedure was repeated across all relevant source–target network pairs: DN-driven (with dATN as target), dATN-driven (with DN as target), and for both positive and negative DN pRFs. For each event type and match status, we computed the average response in the target pRFs at time zero (event peak). These values were entered into a linear mixed effects model with fixed effects of match status (matched vs. anti-matched) and event direction (DN-driven vs. dATN-driven), and with subject as a random effect. This allowed us to test whether retinotopic alignment modulates DN-dATN interactions during spontaneous events and whether such modulation differs for DN-driven vs. dATN-driven events.

Statistical tests

Statistical analyses were implemented in MATLAB (Mathworks, Inc). Given the relatively small number of participants and large number of runs per subject, we used linear mixed effects models to account for both within- and between-subject variability, as described in the relevant sections above. For follow-up tests, we used paired-sample t-tests and corrected for multiple comparisons where appropriate. Corrected p-values are reported where appropriate.